---
title: '
PSYC BC1101 STATISTICS
<a href="syllabus.pdf">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 384 600" style="height:0.75em" aria-labelledby="downloadIconTitle" role="img">
<title id="downloadIconTitle">Download PDF version of syllabus</title>
<!--! Font Awesome Pro 6.2.0 by @fontawesome - https://fontawesome.com License - https://fontawesome.com/license (Commercial License) Copyright 2022 Fonticons, Inc. --><path d="M64 0C28.7 0 0 28.7 0 64V448c0 35.3 28.7 64 64 64H320c35.3 0 64-28.7 64-64V160H256c-17.7 0-32-14.3-32-32V0H64zM256 0V128H384L256 0zM216 232V334.1l31-31c9.4-9.4 24.6-9.4 33.9 0s9.4 24.6 0 33.9l-72 72c-9.4 9.4-24.6 9.4-33.9 0l-72-72c-9.4-9.4-9.4-24.6 0-33.9s24.6-9.4 33.9 0l31 31V232c0-13.3 10.7-24 24-24s24 10.7 24 24z"/></svg></a>
'
---

```{r include=FALSE, eval=FALSE}

# To render a PDF version, comment out the svg download icon in the title and
# run the render command below
quarto::quarto_render("index.qmd", "pdf", "syllabus.pdf")
```


Professor: Dr. Rob Brotherton  
Email: rbrother@barnard.edu

Office Hours: TBD

Lecture hours: TBD 
Lecture venue: TBD




## Course Overview

This course, required for the psychology major, provides an introduction to statistical methods commonly used in psychological research. Topics include measures of central tendency and variability, probability and sampling, confidence intervals and hypothesis testing, t-test and analysis of variance, correlation and regression. In addition to learning the conceptual and mathematical underpinnings of these techniques, you will learn to calculate and interpret statistics with reference to real-world contexts and research questions typical in psychological research. The course includes recitation meetings that provide instruction in the analysis of psychological data using the R statistical software package commonly used by behavioral science researchers.

The course is designed to fulfill these goals:

-   Understand what statistics are, and why and when they are needed
-   Translate observations about the world into statistical statements and questions
-   Learn how statistical methods are used to test psychological hypotheses
-   Use statistical software (R) to describe, visualize, analyze, and report data




## Class Schedule

TBD

```{r eval=FALSE, echo=FALSE}

paste(seq.Date(from = as.Date("2022-01-01"), to = as.Date("2022-05-01"), by = "week"),
      seq.Date(from = as.Date("2022-01-03"), to = as.Date("2022-05-03"), by = "week"), sep = "!") |> 
  stringr::str_split(pattern = "!") |> 
  unlist()


tibble::tribble(
  ~Topic, ~Date,
  "One", "01/01/2022"
)


```



