---
format: 
  revealjs:
    chalkboard: true
---

# 11|THE $t$-TEST

![](covers/11_the-t-test.svg)

```{r cover}
#| eval: false
library(ggplot2)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

dfs <- 1:100
colors <- rep(gg_color_hue(10), 10)

layers <- purrr::map2(.x = dfs,
                      .y = colors,
                     .f = ~stat_function(fun = dt, args = list(df = .x), xlim = c(-3, 3), color = .y))

ggplot() +
  layers +
  scale_x_continuous(expand = c(0,0)) +
  theme_void()

bc1101tools::plot_save("11_the-t-test.svg",
                       subdir = "covers",
                       width = 10, height = 5)

```


# Overview

- [$t$ vs. $z$]
- [The $t$ distribution]
- [The $t$-test]
- [Learning checks]


# $t$ vs. $z$

# $z$-test

- $z$-test
  - Useful if we know everything about original population

![](media/nhst-z-diagram.svg)


## $z$-test problem

- [Problem]{.emph}
  - Often don't know everything about original population

![](media/nhst-z-diagram.svg)

## $t$-test solution

- Estimate population variability using sample

## The $t$ statistic {.smallest}

- Estimated standard error $s_M$ used in place of (unknown) population standard error $\sigma_M$

::: {.c}
$$z = \dfrac{M - \mu}{\require{enclose}\enclose{horizontalstrike}{\sigma_M}}
\ \ \ \ \ \ \ \ \ \ \ \ \ 
t = \dfrac{M - \mu}{s_M}$$

$$
\begin{align}
\text{Standard error} = \sigma_M = \dfrac{\sigma}{\sqrt{n}}
\ \text{or...} \ 
\dfrac{\sqrt{\sigma^2}}{\sqrt{n}} \
\text{or...} \
\sqrt{\dfrac{\sigma^2}{n}}
\\ \\
\text{Estimated standard error} = s_M = \dfrac{s}{\sqrt{n}}
\ \text{or...} \ 
\dfrac{\sqrt{s^2}}{\sqrt{n}} \
\text{or...} \
\sqrt{\dfrac{s^2}{n}}
\end{align}
$$
:::


## Degrees of freedom {.smallest}

- Degrees of freedom
  - $df$ depends on kind of $t$-test you're doing
  - Single sample $t$-test: $df = n - 1$

$\text{Population variance} = \sigma^2 = \dfrac{SS}{N}$

$\text{Sample variance} = s^2 = \dfrac{SS}{df} = \dfrac{SS}{n-1}$

$\text{Sample standard deviation} = s = \sqrt{\dfrac{SS}{df}} = \sqrt{\dfrac{SS}{n-1}}$



# The $t$ distribution {.smaller}

:::: {.columns}

::: {.column width="40%"}
- Degrees of freedom
  - $t$ distribution is actually a family of distributions
  - Slightly different curve for each value of $df$
  - As $df$ increases, $t$ distribution gets closer to normal
  - Because the more degrees of freedom, the better $s^2$ represents $\sigma^2$
:::

::: {.column width="60%"}

```{r t-curves}
#| eval: false
library(bc1101tools)
library(ggplot2)
library(ggrepel)

base <- ggplot() + 
  scale_y_continuous(limits = c(0, .4)) +
  theme_void()

norm <- base + 
  stat_function(fun = dnorm, xlim = c(-3, 3), linetype = 2) +
  geom_label_repel(aes(x = 0, y = .4, label = "Normal distribution"), nudge_x = -1.5, label.size = 0, fill = NA)

norm

t5 <- base + 
  stat_function(fun = dt, args = list(df = 5), xlim = c(-3, 3), color = "steelblue") +
  geom_label_repel(aes(x = 0, y = dt(0, 5), label = "t(df = 5)"), 
                   nudge_x = 1.5, nudge_y = -.01,
                   label.size = 0, color = "steelblue", fill = NA)

t10 <- base + 
  stat_function(fun = dt, args = list(df = 10), xlim = c(-3, 3), color = hsv(.5, 1, .5)) +
  geom_label_repel(aes(x = 0, y = dt(0, 10), label = "t(df = 10)"), nudge_x = 1.5, label.size = 0, color = hsv(.5, 1, .5), fill = NA)

t30 <- base + 
  stat_function(fun = dt, args = list(df = 30), xlim = c(-3, 3), color = "red") +
  geom_label_repel(aes(x = 0, y = dt(0, 30), label = "t(df = 30)"), 
                   nudge_x = 1.5, nudge_y = 1,
                   label.size = 0, color = "red", fill = NA)

w <- 5
h <- 5
plot_save("t-distribution-0.svg", plot = norm, width = w, height = h)
plot_save("t-distribution-1.svg", plot = t5, width = w, height = h)
plot_save("t-distribution-2.svg", plot = t10, width = w, height = h)
plot_save("t-distribution-3.svg", plot = t30, width = w, height = h)
```

::: {.r-stack}

![](media/t-distribution-1.svg){.fragment}

![](media/t-distribution-2.svg){.fragment}

![](media/t-distribution-3.svg){.fragment}

![](media/t-distribution-0.svg){.invertable}
:::
:::

::::


## The $t$ distribution

```{r}

norm_df <- tibble::tibble(value = seq(-4, 4, length.out = 210),
                          density = dnorm(value))

norm_arr <- purrr::transpose(norm_df)


ojs_define(norm_arr)

```

::: {.c}
```{ojs}
#| echo: false

<!-- norm_arr -->

<!-- norm_data -->

jStat = require("https://cdn.jsdelivr.net/npm/jstat@latest/dist/jstat.min.js")

d3 = require("https://d3js.org/d3.v5.min.js")


height = 400
width = 800


viewof df = html`<input type=range min=1 max=50 step=1 value=1 style="width: 50%;">`

<!-- comment -->
tex`df = ${df.toLocaleString("en")}`


data = {
  var values = jStat(-4, 4, 210)[0],
      <!-- df = df, -->
      arr = [];
  for (var i in values) {
    arr.push(
      {
        value: values[i], 
        density: jStat.studentt.pdf(values[i], df)
      }
    )
  }
  return arr;
}

norm_data = {
  var values = jStat(-4, 4, 210)[0],
      <!-- df = df, -->
      arr = [];
  for (var i in values) {
    arr.push(
      {
        value: values[i],
        density: jStat.normal.pdf(values[i], 0, 1)
      }
    )
  }
  return arr;
}

<!-- norm_data -->


chart = {
  const svg = d3.select(DOM.svg(width, height));

  <!-- svg.append("g") -->
  <!--     .call(xAxis); -->

  <!-- svg.append("g") -->
  <!--     .call(yAxis); -->
  
  svg.append("path")
      .datum(data)
      .attr("fill", "none")
      .attr("stroke", "red")
      .attr("stroke-width", 4)
      .attr("stroke-linejoin", "round")
      .attr("stroke-linecap", "round")
      .attr("d", line);
      
  svg.append("path")
      .datum(norm_arr)
      .attr("fill", "none")
      .attr("stroke", "black")
      .attr("stroke-width", 2)
      .attr("stroke-linejoin", "round")
      .attr("stroke-linecap", "round")
      .attr("stroke-dasharray", "5, 5")
      .attr("d", line)
      .attr("class", "invertable");
  
  return svg.node();
}

<!-- margin = ({top: 20, right: 0, bottom: 30, left: 40}) -->
margin = ({top: 20, right: 0, bottom: 0, left: 0})

line = d3.line()
    .x(d => x(d.value))
    .y(d => y(d.density))

x = d3.scaleLinear()
  .domain([d3.min(data, d => d.value * 0.9), d3.max(data, d => d.value * 0.9)]).nice()
  .range([margin.left, width - margin.right])

<!-- y = d3.scaleLinear() -->
<!--   .domain([d3.min(data, d => d.density * 0.9), d3.max(data, d => d.density / 0.9)]) -->
<!--   .range([height - margin.bottom, margin.top]) -->
  
y = d3.scaleLinear()
  .domain([0, 0.4])
  .range([height - margin.bottom, margin.top])

  
  
<!-- xAxis = g => g -->
<!--     .attr("transform", `translate(0,${height - margin.bottom})`) -->
<!--     .call(d3.axisBottom(x) -->
<!--         .tickSizeOuter(0)) -->

<!-- yAxis = g => g -->
<!--     .attr("transform", `translate(${margin.left},0)`) -->
<!--     .call(d3.axisLeft(y)) -->
<!--     .call(g => g.select(".domain").remove()) -->
```
:::

::: {style="font-size: 0.5em; width: 70%; margin-left: 15%"}
[Normal distribution]{style="text-decoration-line: overline; text-decoration-style: dashed;"} [$t$ distribution]{style="color: red; text-decoration-line: overline; float: right;"}
:::


## $t$ table

:::: {.columns}

::: {.column width="50%"}
Table
:::

::: {.column width="50%"}
![](media/directional-test-lower-tail.svg){.invertable width=80%}

![](media/directional-test-upper-tail.svg){.invertable width=80%}

![](media/nondirectional-test.svg){.invertable width=80%}
:::

::::



## $t$ table

:::: {.columns}

::: {.column width="50%"}
Table
:::

::: {.column width="50%"}
- Using R
- `pt()` and `qt()` instead of `pnorm()` and `qnorm()`

```{r}
#| echo: true
#| error: true
qnorm(.05)

qt(.05)

qt(.05, df = 5)

qt(.05, df = 10)
```

:::

::::


# The $t$-test

- Single-sample $t$ test

## Class reaction times

```{r}
#| style: "margin-bottom: 1em;"

library(bc1101tools)

data <- bc1101tools::class_reaction_times
rt <- bc1101tools::spiderman

M <- mean(data, na.rm = TRUE)
SD <- sd(data, na.rm = TRUE)
n <- length(data[!is.na(data)])
sM <- SD/sqrt(n)
mu <- rt$m
t_critical <- qt(.975, df = n - 1)
t <- (M - mu) / sM

print(data)
```

:::: {.columns}

::: {.column width="30%"}
```{r}
#| style: "font-size: 0.5em;"

data <- data[!is.na(data)]

bc1101tools::frequency_table(data, binwidth = 20, values_name = "RT")
```

:::

::: {.column width="70%"}
```{r}
#| eval: false

plot_histogram(data, binwidth = 20, xlab = "Reaction Time (ms)") + bc1101tools::theme_bc1101() +
  ggrepel::geom_label_repel(ggplot2::aes(x = rt$m, y = 0, label = glue::glue("mu = {rt$m}")),
                            nudge_y = 5,
                            color = "red",
                            fill = NA,
                            label.size = 0,
                            segment.linetype = 2)

plot_save("class-reaction-times-histogram.svg",
          width = 6, height = 3)
```

![](media/class-reaction-times-histogram.svg){.invertable}
:::

::::


## Hypothesis test

- Four steps:
  - 1: State the null and alternative hypotheses 
  - 2: Locate the critical region using the $t$ distribution probabilities, $df$, and $\alpha$
  - 3: Calculate the $t$ test statistic
  - 4: Make a decision regarding $H_0$ (null hypothesis)

## 1. State hypotheses

- Step 1: State hypotheses
  - $H_0$: Stats students have the same average reaction times as the general population $\mu = 284$
  - $H_1$: Stats students have different average reaction times to the general population


## 2. Decision criterion

:::: {.columns}

::: {.column width="80%"}
- Specify $\alpha$, identify critical region(s)
- For $t$, depends on $df$ and thus $n$
- For single-sample $t$-test, $df = n – 1$
:::

::: {.column width="20%"}
```{r}
#| eval: false
bc1101tools::t_table(dfs = c(1:6,20:31), qs = .025, dots = c(6, 18))
```

:::

::::

 
  
  
## 3. State hypotheses


- Step 3: Collect data; compute the test statistic
  - Calculate $t$-statistic for the sample mean
  - Quantifies the difference between the observed sample mean and the hypothesized population mean divided by the estimated standard error


:::: {.columns}

::: {.column width="50%"}
$\mu = 284 \\ M = `r round(M, 2)` \\ SD = `r round(SD, 2)` \\ n = `r n`$
:::

::: {.column width="50%"}
$$\begin{align}
t = \dfrac{M - \mu}{s_M} &= \dfrac{`r round(M, 2)` - `r mu`}{`r round(SD, 2)`/\sqrt{`r n`}} \\
&= \dfrac{`r round(M - mu, 2)`}{`r round(sM, 2)`} \\
&= `r round(t, 2)`
\end{align}$$
:::

::::


## 4. State hypotheses

- Step 4a: Make a decision about $H_0$
  - $t = `r round(t, 2)`$  exceeds critical values $[`r round(qt(c(.025, .975), df = n - 1), 2)`]$
  - $p < \alpha$
  - "Statistically significant" difference

# Learning checks


1. What is the difference between the $t$ distribution and the $z$ distribution?
2. The results of a hypothesis test are reported as follows: $t(21) = 2.38, p < .05$.  How many people were in the sample and what was the statistical decision?
3. True/False
<<<<<<< HEAD
    - An effect that exists is more likely to be detected if sample size $n$ is large
    - An effect that exists is less likely to be detected if the sample standard deviation $s$ is large
=======
  - An effect that exists is more likely to be detected if sample size $n$ is large
  - An effect that exists is less likely to be detected if the sample standard deviation $s$ is large
>>>>>>> 7fa4305426aa3b7acab11293edca73746158b7bc
