---
format: 
  revealjs:
    chalkboard: true
---

# [13|INDEPENDENT-SAMPLES $t$-TEST]{.r-fit-text}

![](covers/11_the-t-test.svg)

# Overview

- [Research design]
- [Calculation]
- [Hypothesis test]
- [Learning checks]


# Research design

## Single-sample $t$-test design

- Compare sample against expected population mean based on logic/theory/scale design
- E.g. Happiness
  - Give everyone $10
  - "What is your current level of happiness?"

## Independent-samples design

What if...  


:::: {.columns .c}

::: {.column width="50%"}
Group A:  
Spend this on yourself
:::

::: {.column width="50%"}
Group B:  
Spend this on someone else
:::

::::

- Later... "What is your current level of happiness?"


::: {.footer .reference}
Dunn, E. W., Aknin, L. B., & Norton, M. I. (2014). Prosocial spending and happiness: Using money to benefit others pays off. *Current Directions in Psychological Science, 23*(1), 41-47. https://doi.org/10.1177/0963721413512503
:::


## Independent-samples

```{r}
#| eval: false


bc1101tools::plot_nhst_diagram(text.size = 5,
                               pop1lab = "Partially\nknown\noriginal\npopulation\n(\U03BC)",
                               pop2lab = "Unknown\ntreated\npopulation",
                               sample2lab = "Treated\nSample\n(M, s, n)")

```


```{r}
#| eval: false

bc1101tools::plot_nhst_diagram(text.size = 5,
                               pop1lab = "Unknown\ntreated\npopulation\nA",
                               pop1line = 2, pop2fill = "white",
                               pop2lab = "Unknown\ntreated\npopulation\nB",
                               sample1lab = "Sample A\n(M, s, n)",
                               sample2lab = "Sample B\n(M, s, n)", sample2fill = "white",
                               sample_line = "white")

```

# Calculation {.small}

Basic structure of all $t$-tests:

::: {.c}
$t = \dfrac{ \text{sample statistic} - \text{population parameter} }{\text{estimated standard error}}$
<br><br>
$t = \dfrac{ \text{how different was observed from predicted?} }{\text{how big a difference would we expect by chance?}}$
<br><br>
$t = \dfrac{ \text{data} - \text{hypothesis} }{\text{error}}$
:::


:::: {.columns .c}

::: {.column width="50%"}
Single-sample $t$

$t = \dfrac{M-\mu}{s_M}$
:::

::: {.column width="50%"}
Independent-samples $t$

$t = \dfrac{(M_1-M_2)-(\mu_1-\mu_2)}{s_{M_1-M_2}}$
:::

::::


# Learning checks

1. What values of $t$ would you expect to see if the null hypothesis is true?
2. Which combination of factors is most likely to produce a significant value for an independent-measures $t$ statistic?
  - Small mean difference and large sample variances
  - Small mean difference and small sample variances
  - Large mean difference and large sample variances
  - Large mean difference and small sample variances
