---
format: revealjs
---

# [13|INDEPENDENT-SAMPLES $t$-TEST]{.r-fit-text}

::: {#cover}
:::

<!-- ![](covers/13_independent-samples-t-test.svg) -->

# Overview

- [Research design]
- [Calculation]
- [Hypothesis test]
- [Learning checks]


# Research design

## Single-sample $t$-test design

- Compare sample against expected population mean based on logic/theory/scale design
- E.g. give everyone $10

:::: {.columns}

::: {.column width="60%"}

::: {#likert-container .smallest style="width: 80%; display: block;"}

What is your current level of happiness?

1. A lot less than usual
2. A little less than usual
3. About average
4. A little more than usual
5. A lot more than usual

:::

:::

::: {.column width="40%" .center-element}

$\mu = 3$

![](media/theoretical-distribution.svg){.invertable style="margin-top: 0;"}
:::

::::

```{ojs}
import {likert} from "../ojs/utils.qmd";
```


## Single-sample $t$-test logic

::: {style="width: 60%; position: relative; font-size: 0.7em; margin-top: 2em; margin-left: 50%; transform: translateX(-50%);"}

:::: {#nhst-diagram}
::: {#lines}
:::
::: {#treatment}
:::
::: {#circle style="border: 2px dashed"}
Partially known  
original population    
$\mu$
:::
::: {#circle style="right: 0; background-color: lightblue; border: 2px dashed"}
Unknown  
treated  
population
:::
::: {#rect}
Sample
:::
::: {#rect style="right: 0; background-color: lightblue;"}
Treated sample 
$n, M, SD$
:::
::: {#population-comparison-line}
:::
::::
:::



## Independent-samples design {.small}

What if... give everyone $10


:::: {.columns .c}

::: {.column width="50%"}
Group A:  
Spend this on [yourself]{.emph}

::: {#likert-container .smallest style="width: 80%; display: block; text-align: left;"}

What is your current level of happiness?

1. A lot less than usual
2. A little less than usual
3. About average
4. A little more than usual
5. A lot more than usual

:::
:::

::: {.column width="50%"}
Group B:  
Spend this on [someone else]{.emph}

::: {#likert-container .smallest style="width: 80%; display: block; text-align: left;"}

What is your current level of happiness?

1. A lot less than usual
2. A little less than usual
3. About average
4. A little more than usual
5. A lot more than usual

:::
:::

::::

::: {.reference}
Dunn, E. W., Aknin, L. B., & Norton, M. I. (2014). Prosocial spending and happiness: Using money to benefit others pays off. *Current Directions in Psychological Science, 23(1)*, 41-47. [https://doi.org/10.1177/0963721413512503](https://doi.org/10.1177/0963721413512503)
:::



## Independent-samples $t$-test logic

::: {style="width: 60%; position: relative; font-size: 0.7em; margin-top: 2em; margin-left: 50%; transform: translateX(-50%);"}

:::: {#nhst-diagram}
::: {#lines}
:::
::: {#treatment}
:::
::: {#circle style="border: 2px dashed"}
Unknown  
treated population  
A
:::
::: {#circle style="right: 0; border: 2px dashed;"}
Unknown  
treated population  
B
:::
::: {#rect style="background-color: lightblue; color: var(--text-color-light);"}
Sample A  
$n, M, SD$
:::
::: {#rect style="right: 0; background-color: lightblue; color: var(--text-color-light);"}
Sample B  
$n, M, SD$
:::
::: {#population-comparison-line}
:::
::::
:::


# Calculation {.small}

Basic structure of all $t$-tests:

::: {.c}
$t = \dfrac{ \text{sample statistic} - \text{population parameter} }{\text{estimated standard error}}$
<br><br>
$t = \dfrac{ \text{how different was observed from predicted?} }{\text{how big a difference would we expect by chance?}}$
<br><br>
$t = \dfrac{ \text{data} - \text{hypothesis} }{\text{error}}$
:::


:::: {.columns .c}

::: {.column width="50%"}
Single-sample $t$

$t = \dfrac{M-\mu}{s_M}$
:::

::: {.column width="50%"}
Independent-samples $t$

$t = \dfrac{(M_1-M_2)-(\mu_1-\mu_2)}{s_{M_1-M_2}}$
:::

::::


## Equations {.small}

:::: {.columns}
::: {.column width="50%"}
- Denominator: $s_{M_1-M_2}$
  - Estimated standard error of the mean difference<br><br>
:::

::: {.column width="50%"}
$s_{M_1-M_2} = \sqrt{\dfrac{s_p^2}{n_1}+\dfrac{s_p^2}{n_2}}$
:::
::::

:::: {.columns}
::: {.column width="50%"}
- $s_p^2$: Pooled variance
  - Weighted average of two sample variances
:::

::: {.column width="50%" style="margin-left: -4em;"}
$\begin{align} s_p^2 = &\dfrac{SS_1+SS_2}{df_1+df_2} \\ \\ \text{or... } &\dfrac{df_1*s_1^2 + df_2*s_2^2}{df_1+df_2} \\ \text{because... } s^2 = &\dfrac{SS}{df} \text{ so... } SS = df*s^2 \end{align}$
:::
::::

## Calculating independent samples $t$

1. Pooled variance: $s_p^2 = \dfrac{SS_1+SS_2}{df_1+df_2}$
2. Estimated standard error of mean difference: $s_{M_1-M_2} = \sqrt{\dfrac{s_p^2}{n_1}+\dfrac{s_p^2}{n_2}}$
3. $t$ statistic: $t = \dfrac{(M_1-M_2)-(\mu_1-\mu_2)}{s_{M_1-M_2}}$


# Hypothesis test

## Step 1: State hypotheses

- Null: There is no difference between groups
  - The treatment has no effect
  - $μ_1 – μ_2 = 0$
- Alternative: There is a difference
  - The treatment has an effect
  - Directional: $μ_1 – μ_2 < 0$ or $μ_1 – μ_2 > 0$
  - Nondirectional: $μ_1 – μ_2 \ne 0$

## Step 2: Define critical region

:::: {.columns}

::: {.column width="50%"}
- Depends on $\alpha$ and $df$

$$
\begin{align}
df &= df_1 + df_2 \\
&= (n_1 – 1) + (n_2 – 1) \\
&= N - 2
\end{align}
$$
![](media/nondirectional-test-labelled.svg)

:::

::: {.column width="50%"}
```{r}
#| style: "font-size: 0.4em;"
bc1101tools::t_table(alphas = .05, two_tailed_only = TRUE)
```

:::

::::



## Step 3: Calculate $t$ statistic

```{r data}
conditionA <- c(1, 5, 2, 4, 3)
conditionB <- c(5, 5, 2, 5, 3)
r <- function(x) {round(x, 2)}
# t.test(conditionA, conditionB, var.equal=T, paired=F)

SS <- function(x) sum((x - mean(x))^2)

m1 <- mean(conditionA)
m2 <- mean(conditionB)
n1 <- length(conditionA)
n2 <- length(conditionB)
N <- n1 + n2
df1 <- n1 - 1
df2 <- n2 - 1
ss1 <- SS(conditionA)
ss2 <- SS(conditionB)
sd1 <- sd(conditionA)
sd2 <- sd(conditionB)

s2p <- (ss1 + ss2)/(df1 + df2)

sm1m2 <- sqrt((s2p/n1)+(s2p/n2))

t <- (m1-m2)/sm1m2
```

:::: {.columns .center-element style="font-size: 0.7em;"}

::: {.column width="50%"}
Spend $10 on self

```{r}
bc1101tools::SD_table(conditionA)
```
:::

::: {.column width="50%"}
Spend $10 on other

```{r}
bc1101tools::SD_table(conditionB)
```
:::

::::



## Step 3 : Calculate $t$ statistic

::: {.center-element}
$s_p^2 = \dfrac{SS_1+SS_2}{df_1+df_2} = \dfrac{`r ss1` + `r ss2`}{`r df1` + `r df2`} = `r s2p`$

<br>

$s_{M_1-M_2} = \sqrt{\dfrac{s_p^2}{n_1}+\dfrac{s_p^2}{n_2}} = \sqrt{\dfrac{`r s2p`}{`r n1`}+\dfrac{`r s2p`}{`r n2`}} = `r round(sm1m2, 2)`$

<br>

$t = \dfrac{(M_1-M_2)-(\mu_1-\mu_2)}{s_{M_1-M_2}} = \dfrac{`r m1` - `r m2`}{`r round(sm1m2, 2)`} = `r round(t, 2)`$
:::


## Step 4: Make decision


## Step 4b: Effect size
  - Not required for nonsignificant differences

## Step 5: Report results

“A two-tailed independent-samples $t$ test suggested that the difference in average happiness between people in the "Spend on self group" $(M = `r m1`$; $SD = `r r(sd1)`)$ and the "Spend on other" group $(M = `r m2`$; $SD = `r r(sd2)`)$ was nonsignificant; $t(`r N-2`) =$ $`r r(t)`$, $p > .05$.”


# Assumptions

Assumptions for independent-samples $t$-tests

1. The observations within each sample must be independent
2. The two populations from which the samples are selected must be normal
    - Can be ignored with large enough sample size
3. The two populations from which the samples are selected must have equal variances
    - Homogeneity of variance 
    - Because pooled variance is a weighted average

## Homogeneity of variance

- Testing the homogeneity of variance assumption
  - Hartley’s F-max test

::: {.center-element}
$F_{max} = \dfrac{s^2_{largest}}{s^2_{smallest}}$
:::

- Small value (near 1) indicates similar sample variances, larger values indicate larger difference
- Look up associated critical value for $F$-max test
- If value exceeds critical value, indicates homogeneity assumption has been violated

## Homogeneity of variance correction

- If homogeneity of variance assumption is violated...
  - Calculate standard error without pooled variance
  - Adjust $df$ using equation:

::: {.center-element}
$$df = \dfrac{(\dfrac{s_1^2}{n_1}+\dfrac{s_2^2}{n_2})}
{\dfrac{(\dfrac{s_1^2}{n_1})^2}{n_1-1} + \dfrac{(\dfrac{s_2^2}{n_2})^2}{n_2-1}
}$$
:::

## Homogeneity of variance correction

- …or let R do the work for you
    - `t.test()` function automatically applies correction
    - Specify `var.equal = TRUE` to override


:::: {.columns .small}

::: {.column width="50%"}
```{r}
#| echo: true

conditionA <- c(1, 5, 2, 4, 3)
conditionB <- c(5, 5, 2, 5, 3)

t.test(x = conditionA, y = conditionB)

```
:::

::: {.column width="50%"}
```{r}
#| echo: true

conditionA <- c(1, 5, 2, 4, 3)
conditionB <- c(5, 5, 2, 5, 3)

t.test(x = conditionA, y = conditionB, 
       var.equal = TRUE)

```
:::

::::

# Confidence interval

- Quantifies precision of our statistic
  - Distribution of differences between two samples drawn from a population with a given standard error of the mean difference

::: {.center-element}
$t = \dfrac{(M_1-M_2)-(\mu_1-\mu_2)}{s_{M_1-M_2}}$

$(\mu_1-\mu_2) = (M_1-M_2) \pm t * s_{M_1-M_2}$
:::

# Learning checks

1. What values of $t$ would you expect to see if the null hypothesis is true?
2. Which combination of factors is most likely to produce a significant value for an independent-measures $t$ statistic?
  - Small mean difference and large sample variances
  - Small mean difference and small sample variances
  - Large mean difference and large sample variances
  - Large mean difference and small sample variances



::: {.content-hidden}

# Data and figures

## Cover

```{r cover-data}

# library(ggplot2)
library(dplyr)
n <- 5000
set.seed(1)
data <- tibble::tibble(group = sample(c("A","B"), n, replace = T),
                       x = rnorm(n, mean = 100, sd = 19) %>% round(0),
                       y = 1) |> 
  mutate(x = ifelse(group=="B", x+15, x),
         n = 1:n()) |> 
  group_by(x, group) |> 
  mutate(y_cum = cumsum(y)) |> 
  ungroup() %>%
  arrange(group)  |> 
  group_by(x, y_cum) |> 
  mutate(g = n(),
         group = ifelse(g==2, "AB", group))

data <- purrr::transpose(rename(data, value = x))
ojs_define(data)
```


```{r cover-ggplot}
#| eval: false

p <- ggplot(data, aes(x = x, y = y_cum, fill = group)) +
  ggforce::geom_regon(aes(x0 = x + .1, y0 = y_cum-.5, sides = 4, r = sqrt((.5^2)*2)*.85, angle = 0), color = "red", alpha = .4) +
#  geom_bar(stat = "identity", fill = NA, color = "green") +
  coord_fixed() +
  scale_x_continuous(limits = c(40,180), expand = c(0,0)) +
  scale_fill_manual(values = c("firebrick","dodgerblue")) +
#  transition_reveal(n) +
  theme_void() +
  theme(legend.position = 'none',
        panel.background = element_blank()) +
  NULL
p
bc1101tools::plot_save("independent-samples-t-test.svg", subdir = "covers", width = 10, height = 5)

related <- p + scale_fill_manual(values = c("dodgerblue","dodgerblue3"))
related
```

:::

```{ojs}
data
cover = {
  const w = 1050
  const h = 500
  const x = d3.scaleLinear()
    .domain([40,180])
    .range([0,w])
  const y = d3.scaleLinear()
    .domain([0,70])
    .range([h,0])
  const fill = d3.scaleOrdinal()
    .domain(["A","B", "AB"])
    .range(["#dda6a6", "#b0d3ff", "#9c9ecb"])
    
  const svg = d3.select("#cover").append("svg").attr("width", w).attr("height", h)
  
  svg.selectAll("rect").data(data).enter().append("rect")
    .attr("x", d => x(d.value))
    .attr("y", d => y(d.y_cum))
    .attr("width", 6)
    .attr("height", 6)
    .style("fill", d => fill(d.group))
}
```


::: {.content-hidden}
## NHST diagrams
```{r}
#| eval: false
library(bc1101tools)

bc1101tools::plot_nhst_diagram(text.size = 5,
                               pop1lab = "Partially\nknown\noriginal\npopulation\n(\U03BC)",
                               pop2lab = "Unknown\ntreated\npopulation",
                               sample2lab = "Treated\nSample\n(M, s, n)")

plot_save("nhst-diagram-single-sample-t.svg", width = 5, height = 4)

bc1101tools::plot_nhst_diagram(text.size = 5,
                               pop1lab = "Unknown\ntreated\npopulation\nA",
                               pop1line = 2, pop2fill = NA,
                               pop2lab = "Unknown\ntreated\npopulation\nB",
                               sample1lab = "Sample A\n(M, s, n)",
                               sample2lab = "Sample B\n(M, s, n)", sample2fill = NA,
                               sample_line = NA)

plot_save("nhst-diagram-independent-samples-t.svg", width = 5, height = 4)

```

:::
