---
format: revealjs
---

# 16|ANOVA [part 2]{style="float: right; opacity: 0.8;"}

![](covers/15_ANOVA.svg)

# Overview

- [ANOVA terminology]
- [Calculating ANOVA]
- [Hypothesis test]
- [Post-hoc tests]
- [Assumptions]
- [Learning checks]


# ANOVA terminology

```{r}

x <- c(9,11,13)
y <- c(3,5,4)
z <- c(5,6,7)

table2 <- knitr::kable(data.frame(x, y, z), col.names = c("🍌<br>Banana", "🍬<br>Candy", "😐<br>Control"), align = "c", format = "html", escape = FALSE) |> 
  kableExtra::add_header_above(c("Manipulation" = 3)) |> 
  kableExtra::column_spec(1:3, width = "5em")

means <- data.frame(glue::glue('\\(M = {mean(x)}\\)'),
                    glue::glue('\\(M = {mean(y)}\\)'),
                    glue::glue('\\(M = {mean(z)}\\)'))

table2_footer <- knitr::kable(means, col.names = NULL, align = "c") |> 
  kableExtra::column_spec(1:3, width = "5em")
```


:::: {.columns}

::: {.column width="50%" style="font-size: 0.8em;"}
- Factor
  - The independent variable
  - Designates the groups being compared
  - Between-participants / within-participants
- Levels
  - Conditions that make up a factor
- E.g.
  - "1x3 between-participants ANOVA"
:::

::: {.column width="50%"}
```{r}
#| style: "font-size: 0.65em; margin-top: 1em; margin-bottom: -1em;"
table2
table2_footer
```

![](media/anova-populations-3.svg){.invertable width="80%"}
:::

::::



# Calculating ANOVA {.smaller}

- Step 1: Calculate $SS$ and $df$ (total, within, between)

<style>
  .xx-total {
    display: table;
    margin: 0 auto;
    margin-bottom: 2em;
  }
  .xx-total:before {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(135deg) scale(0.5);
  }
  .xx-total:after {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(45deg) scale(0.5);
  }
  .xx-between {
    float: right;
    margin-right: 20%;
    margin-bottom: 1em;
  }
  .xx-within {
    float: left;
    margin-left: 20%;
  }
</style>

:::: {.columns}

::: {.column width="50%"}

[$SS_{total}$]{.xx-total}

[$SS_{within}$]{.xx-within} [$SS_{between}$]{.xx-between}

:::

::: {.column width="50%"}
[$df_{total}$]{.xx-total}

[$df_{within}$]{.xx-within} [$df_{between}$]{.xx-between}
:::

::::

- Step 2: Calculate variances (Mean Squares)

:::: {.columns .center-element}

::: {.column width="50%"}
$MS_{between} = \dfrac{SS_{between}}{df_{between}}$
:::

::: {.column width="50%"}
$MS_{within} = \dfrac{SS_{within}}{df_{within}}$
:::

::::

- Step 3: Calculate $F$-ratio

::: {.center-element}
$F = \dfrac{MS_{between}}{MS_{within}}$
:::

## $SS$ and $df$

:::: {.columns}

::: {.column width="50%"}
$SS_{total} = \Sigma X^2 - \dfrac{G^2}{N}$
$df_{total} = N-1$

$SS_{within} = \Sigma SS_{each \ treamtent}$
$df_{within} = N-k$

$SS_{between} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N}$
$df_{between} = k-1$
:::

::: {.column width="50%"}
```{r}
#| style: "font-size: 0.8em;"

anova_terms <- data.frame(Symbol = c("\\(k\\)", 
                                     "\\(n_1, n_2...\\)",
                                     "\\(N\\)",
                                     "\\(T_1, T_2...\\)",
                                     "\\(G\\)"),
                          Meaning = c("Number of treatment conditions",
                                      "Number of scores in each treatment",
                                      "Total number of scores",
                                      "Sum of scores \\((\\Sigma X)\\) for each treatment",
                                      "Grand total of all scores in the study"))

knitr::kable(anova_terms, format = 'html', escape=F)
```

:::

::::


## Summary table

```{r}
s_tab <- data.frame(Source = c("Between treatments","Within treatments","Total"),
                    SS = c("","",""),
                    df = c("","",""),
                    MS = c("","",""),
                    F = c("","",""))

knitr::kable(s_tab, format = 'html', 
             col.names = c("Source", "\\(SS\\)", "\\(df\\)", "\\(MS\\)", "\\(F\\)"))
```


# Hypothesis test

## Step 1. State hypotheses

- $H_0: \mu_1 = \mu_2 = \mu_3$
  - No treatment effect
  - Numerator & denominator should be about the same
  - $F$ should be near $1.00$
- $H_1$ : At least one population mean differs from another
  - There is some treatment effect
  - Numerator bigger than denominator
  - $F$ should be noticeably larger than $1.00$

## Step 2. Critical region

- Like $t$ distributions, there is a different $F$ distribution for each value of $df$
  - Now we have two different $df$ values
  - $df$ numerator $(df_{between})$ 
  - $df$ denominator $(df_{within})$
  - Note, distribution isn’t symmetrical
  - $F$ values are always positive

## $F$ table

::: {#tbl}
```{r}
#| style: "font-size: 0.5em;"
options(digits = 3)

dfs <- setNames(1:10, 1:10)

aov_tab <- purrr::map_df(.x = dfs, ~qf(.95, df1 = 1:10, df2 = .x)) |> 
  t() |>
  as.data.frame() |>
  setNames(1:10) |>
  tibble::rownames_to_column(var = "df_denominator")


knitr::kable(aov_tab, format = "html", escape=FALSE, align = "c", digits = 3,
               col.names = c("\\(df_{denominator}\\)", 1:10)) |>
  kableExtra::add_header_above(c("\\(\\alpha = .05\\)", "\\(df_{numerator}\\)" = 10), escape=FALSE, align = "left")

```
:::


```{ojs}

jStat = require("../js/jstat.js")

aov_table = {

  var df1, df2, critical
  
  function getDf2() {
    try {
      df2 = d3.select(this).select("td")._groups[0][0].innerHTML;
    } catch {} finally {update();}
  }
  
    function getDf1() {
    try {
      df1 = d3.select(this)._groups[0][0].cellIndex
      if (df1 < 1) {df1 = 1}
    } catch {} finally {update();}
  }
  
  d3.select("#tbl").selectAll("tr").on("click", getDf2)
  d3.select("#tbl").selectAll("td").on("click", getDf1)


  const w = 1050
  const h = 200
  const margin = {right: 100, left: 100, bottom: 20}
  
  const svg = d3.select("#tbl").append("svg")
    .attr("width", w).attr("height", h)
    

  const x = d3.scaleLinear()
    .domain([0,4])
    .range([margin.left, w - margin.right])
  const y = d3.scaleLinear()
    .domain([0,1])
    .range([h - margin.bottom, 0])
  const xAxis = d3.axisBottom(x)
  const line = d3.line()
    .x(d => x(d.value))
    .y(d => y(d.density))
  
  function makeCurve(df1, df2) {
    var arr = []
    var values = jStat(0.01, 4, 210)[0]
    for (var i = 0; i < values.length; i++) {
      arr.push({value: values[i],
                density: jStat.centralF.pdf(values[i], Number(df1), Number(df2))})
    }
    return arr
  }
  
  svg.append("g").attr("transform", `translate(0, ${h - margin.bottom})`)
    .call(xAxis)
    
  const curve = svg.append("path")
    .attr("d", line(makeCurve(10, 10)))
    .style("fill", "none")
    .style("stroke", "black")
  
  const df1Input = document.getElementById('df1-input')
  df1Input.oninput = function() {
    df1 = df1Input.value
    update();
  }
  
  const df2Input = document.getElementById('df2-input')
  df2Input.oninput = function() {
    df2 = df2Input.value
    update();
  }
  
  function update() {
    d3.select("#df1-value").text(df1)
    d3.select("#df2-value").text(df2)
    critical = jStat.centralF.inv(0.95, Number(df1), Number(df2))
    curve.attr("d", line(makeCurve(Number(df1), Number(df2))))
  }
  


  
}


```

::: {#controls style="position: absolute; top: 75%; left: 50%; font-size: 0.8em; float: right;"}
<label for="df1-input">\\(df_{numerator} = \\)</label><span id="df1-value" class="math" style="display: inline-block; width: 1em;">1</span><input data-prevent-swipe type="range" id="df1-input" name="df1" value="1" min="1" max="10" style="border: none; height: 1em; margin-left: 0.5em; width: fit-content; vertical-align: unset;">

<label for="df2-input">\\(df_{denominator} = \\)</label><span id="df2-value" class="math" style="display: inline-block; width: 1em;">1</span><input data-prevent-swipe type="range" id="df2-input" name="df2" value="1" min="1" max="100" style="border: none; height: 1em; font-family: KaTeX_Main; font-size: 1em; margin-left: 0.5em; width: fit-content; vertical-align: unset;">
:::

## Step 3. Calculate test statistic

```{r data-setup}
options(digits = 2)

x <- c(9,11,13)
y <- c(3,5,4)
z <- c(5,6,7)

a2 <- bc1101tools::example_ANOVA(x, y, z)

N <- length(c(x, y, z))
n <- length(x)
k <- 3
G <- sum(c(x, y, z)) 
T1 <- sum(x)
T2 <- sum(y)
T3 <- sum(z)
sigma_x_sq <- sum(c(x,y,z)^2)
eta_sq <- a2$SS_between / a2$SS_total

```

:::: {.columns}

::: {.column width="50%"}
```{r}
#| style: "font-size: 0.8em"

table2 <- knitr::kable(data.frame(x, y, z), col.names = c("🍌<br>Banana", "🍬<br>Candy", "😐<br>Control"), align = "c", format = "html", escape = FALSE) |> 
  kableExtra::add_header_above(c("Manipulation" = 3)) |> 
  kableExtra::column_spec(1:3, width = "5em")

means <- data.frame(glue::glue('\\(M = {mean(x)}\\)'),
                    glue::glue('\\(M = {mean(y)}\\)'),
                    glue::glue('\\(M = {mean(z)}\\)'))

table2_footer <- knitr::kable(means, col.names = NULL, align = "c") |> 
  kableExtra::column_spec(1:3, width = "5em")


table2
table2_footer
```

:::

::: {.column width="50%"}
$N = `r N`$  
$n = `r n`$  
$k = `r k`$  
$G = `r G`$  
$\Sigma X^2 = `r sigma_x_sq`$
:::

::::


## Step 3. $SS$ and $df$s

$SS_{total} = \Sigma X^2 - \dfrac{G^2}{N} = `r sigma_x_sq` - \dfrac{`r G`^2}{`r N`} = `r a2$SS_total`$

$SS_{within} = \Sigma SS_{each \ treamtent} = 8+2+2 = `r a2$SS_within`$

$SS_{between} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N} = \dfrac{`r T1`^2}{`r n`}+\dfrac{`r T2`^2}{`r n`}+\dfrac{`r T3`^2}{`r n`} - \dfrac{`r G`^2}{`r N`}  = `r a2$SS_between`$

$df_{total} = N-1 = `r a2$df_total`$

$df_{within} = N-k = `r a2$df_within`$

$df_{between} = k-1 = `r a2$df_between`$


## Step 3. $MS$ and $F$

$MS_{between} = \dfrac{SS_{between}}{df_{between}} = `r a2$MS_between`$

$MS_{within} = \dfrac{SS_{within}}{df_{within}} = `r a2$MS_within`$


$F = \dfrac{MS_{between}}{MS_{within}} = \dfrac{`r a2$MS_between`}{`r a2$MS_within`} = `r a2$f_ratio`$

## Step 4. Make decision

- $F > F_{critical}$?
  - Reject or fail to reject $H_0$

- Step 4b. Effect size
  - Compute percentage of variance accounted for by treatment
  - $r^2$ concept (proportion of variance explained)
  - For ANOVA called $\eta^2$ (“eta squared”)

::: {.center-element}
$\eta^2 = \dfrac{SS_{between}}{SS_{total}} = `r eta_sq`$
:::

## Report results {.small}

- Descriptives
  - Treatment means and standard deviations are presented in text, table and/or graph
- Hypothesis test outcome
  - Results of ANOVA are summarized, including
  - $F$ and $df$ values, $p$, $\eta^2$ (if significant)

> A single-factor, independent-samples ANOVA revealed a significant difference between people who consumed a banana ($M = `r a2$mean_x`$; $SD = `r sd(x)`$), a candy bar ($M = `r a2$mean_y`$; $SD = `r sd(y)`$), and the control condition ($M = `r a2$mean_z`$; $SD = `r sd(z)`$); $F(`r a2$df_between`,`r a2$df_within`) = `r a2$f_ratio`$, $p < .05$, $\eta^2 = `r eta_sq`$.

# Post-hoc tests

- Post hoc tests compare two means at a time 
  - [Pairwise]{.emph} comparisons
  - Each comparison includes risk of a Type I error
  - Risk of Type I error accumulates 
  - Experimentwise alpha level, $\alpha_{experimentwise}$
- Post-hoc tests use special methods to control experimentwise Type I error rate

## Tukey's $HSD$

::: {.smallest}
- Tukey’s Honestly Significant Difference
  - Minimum difference between pairs of treatment means so that $p < \alpha_{experimentwise}$
  - $q$ is the Studentized Range statistic
  - Depends on $\alpha$, $k$, and $df$ for denominator
  - Find $q$ in table or R
:::

:::: {.columns}

::: {.column width="40%"}

$$\begin{align}
HSD &= q \sqrt{\dfrac{MS_{within}}{n}} \\
    &= 4.34 \sqrt{\dfrac{2}{3}} \\
    &= 3.54
\end{align}$$

:::

::: {.column width="60%"}

```{r}
#| style: "font-size: 0.6em;"

options(digits = 3)

tukey_df <- rbind(
  qtukey(.05, 2:6, 5, lower.tail = FALSE),
  qtukey(.05, 2:6, 6, lower.tail = FALSE),
  qtukey(.05, 2:6, 7, lower.tail = FALSE),
  qtukey(.05, 2:6, 8, lower.tail = FALSE),
  qtukey(.05, 2:6, 9, lower.tail = FALSE),
  qtukey(.05, 2:6, 10, lower.tail = FALSE)) |> 
  as.data.frame() |> 
  tibble::add_column(df = c(5,6,7,8,9,10), .before = 1)

knitr::kable(tukey_df, digits = 2, col.names = c(" ", 2:6)) |> 
  kableExtra::add_header_above(c("\\(df\\)" = 1, "Number of Conditions" = 5)) |> 
  kableExtra::column_spec(1, bold = TRUE)

```

:::

::::


## Tukey's $HSD$

::: {.smallest}
- Tukey’s Honestly Significant Difference
  - Minimum difference between pairs of treatment means so that $p < \alpha_{experimentwise}$
  - $q$ is the Studentized Range statistic
  - Depends on $\alpha$, $k$, and $df$ for denominator
  - Find $q$ in table or R
:::

:::: {.columns}

::: {.column width="40%"}

$$\begin{align}
HSD &= q \sqrt{\dfrac{MS_{within}}{n}} \\
    &= 4.34 \sqrt{\dfrac{2}{3}} \\
    &= 3.54
\end{align}$$

:::

::: {.column width="60%"}

```{r}
#| style: "font-size: 0.65em; margin-top: 1em; margin-bottom: -1em;"
table2
table2_footer
```

:::

::::

# Assumptions

1. Observations within each sample must be independent
2. Population from which the samples are selected must be normal
3. Populations from which the samples are selected must have equal variances (homogeneity of variance)

# Learning checks

- For independent-samples (between-participants) ANOVA, what do the following represent?
  - $N$
  - $n_1 ,n_2 , n_3$ , etc
  - $G$
  - $T$
  - $MS_{between}$
  - $MS_{within}$



