---
format: revealjs
---

# [17|RELATED-SAMPLES ANOVA]{.r-fit-text}

![](covers/17_related-samples-ANOVA.svg)


# Overview

- [Logic]
- [Calculations]
- [Post-hoc tests]
- [Learning checks]


# Logic

## Independent-samples

:::: {.columns}

::: {.column width="50%"}
- Independent-measures ANOVA
  - "Between-participants"
  - Different participants in each treatment
  - Between-groups variability could be due to treatment effect and/or individual differences
:::

::: {.column width="50%"}

::: {#independent-samples-container}
:::

:::

::::


## Related samples


:::: {.columns}

::: {.column width="50%"}
- Repeated-measures (within-participants)
  - Same participants in each treatment
- Or matched-samples
  - Matched participants in each treatment
:::

::: {.column width="50%"}

::: {#related-samples-container}
:::

:::

::::


## Related samples


:::: {.columns}

::: {.column width="50%" style="font-size: 0.8em;"}
- Advantages
  - Eliminates individual differences as a source of variability between treatments
  - Smaller number of participants needed 
- Disadvantages
  - Something other than the treatment may cause participant’s scores to change
  - E.g. practice, world events, natural improvement
:::

::: {.column width="50%"}

::: {#related-samples-container}
:::

:::

::::


## Partitioning variance

- Independent samples

:::: {#diagram-container style="position: relative; width: 80%; height: 300px; font-size: 0.6em; margin: 0 auto;"}

::: {.node .top .total}
Total variance
:::
::: {.node .between style="top: 50%; left: 0"}
Variance between treatments  
$MS_{between}$
:::
::: {.node .text style="top: 75%; left: 0; z-index: -1;"}
- Treatment effect
- Sampling error
- Individual differences
:::
::: {.node .within style="top: 50%; right: 0;"}
Variance within groups  
$MS_{within}$
:::
::: {.node .text style="top: 75%; right: 0; z-index: -1;"}
- Sampling error
- Individual differences
:::

::::


::: {.center-element style="margin-top: 3em;"}
$F = \dfrac{{MS_{between}}}{{MS_{within}}} = \dfrac{treatment \cdot error \cdot ind. \ diff.}{error \cdot ind. \ diff.}$
:::


## Partitioning variance

- Related samples: problem

:::: {#diagram-container style="position: relative; width: 80%; height: 300px; font-size: 0.6em; margin: 0 auto;"}

::: {.node .top .total}
Total variance
:::
::: {.node .between style="top: 50%; left: 0"}
Variance between treatments   
$MS_{between \ treatments}$
:::
::: {.node .text style="top: 75%; left: 0; z-index: -1;"}
- Treatment effect
- Sampling error
:::
::: {.node .within style="top: 50%; right: 0;"}
Variance within groups  
$MS_{within}$
:::
::: {.node .text style="top: 75%; right: 0; z-index: -1;"}
- Sampling error
- Individual differences
:::

::::


::: {.center-element style="margin-top: 3em;"}
$F = \dfrac{{MS_{between \ treatments}}}{{MS_{within}}} = \dfrac{treatment \cdot error}{error \cdot ind. \ diff.}$
:::


## Partitioning variance

- Related samples: solution

:::: {#diagram-container style="position: relative; width: 80%; height: 300px; font-size: 0.6em; margin: 0 auto;"}

::: {.node .top .total}
Total variance
:::
::: {.node .between style="top: 50%; left: 8%"}
Variance between treatments   
$MS_{between \ treatments}$
:::
::: {.node .text style="top: 75%; left: 8%; z-index: -1;"}
- Treatment effect
- Sampling error
:::
::: {.node .within style="top: 50%; right: 8%;"}
Variance within groups  
$MS_{within}$
:::
::: {.node .error style="top: 85%; right: 30%; width: 25%;"}
Error  
$MS_{error}$
:::
::: {.node .btwn-ss style="top: 85%; right: 0; width: 25%"}
Between subjects  
$MS_{between \ S's}$
:::
::: {.node .text style="top: 110%; right: 30%; z-index: -1; width: 25%;"}
- Sampling error
:::
::: {.node .text style="top: 110%; right: 0; z-index: -1; width: 25%;"}
- Individual differences
:::

::::


::: {.center-element style="margin-top: 3em;"}
$F = \dfrac{{MS_{between \ treatments}}}{{MS_{error}}} = \dfrac{treatment \cdot error}{error}$
:::

# Calculations

## Calculations: $SS$ and $df$

<style>
  .xx-total {
    display: table;
    margin: 0 auto;
    margin-bottom: 2em;
  }
  .xx-total:before {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(135deg) scale(0.5);
  }
  .xx-total:after {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(45deg) scale(0.5);
  }
  .xx-between {
    float: left;
    margin-left: 20%;
    margin-right: -4em;
    margin-bottom: 2em;
  }
  .xx-within {
    float: right;
    margin-right: 20%;
    <!-- margin-top: -2.5em; -->
    <!-- margin-bottom: 4em; -->
  }
  .xx-within:before {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(135deg) scale(0.5);
  }
  .xx-within:after {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(45deg) scale(0.5);
  }
  .xx-error {
    float: left;
    margin-left: 3em;
    margin-top: 2em;
  }
  .xx-btwn-ss {
    float: right;
    margin-top: 2em;
  }
</style>

:::: {.columns}

::: {.column width="50%"}

[$SS_{total}$]{.xx-total}

[$SS_{btwn \ T's}$]{.xx-between} [$SS_{within}$]{.xx-within}

[$SS_{error}$]{.xx-error} [$SS_{btwn \ S's}$]{.xx-btwn-ss}
:::

::: {.column width="50%"}
[$df_{total}$]{.xx-total}

[$df_{btwn \ T's}$]{.xx-between} [$df_{within}$]{.xx-within}

[$df_{error}$]{.xx-error} [$df_{btwn \ S's}$]{.xx-btwn-ss}
:::

::::

:::: {.columns}

::: {.column .smallest}

$$
SS_{total} = \Sigma X^2 - \dfrac{G^2}{N} \\
SS_{within} = \Sigma SS_{within \ each \ treatment} \\
SS_{between \ treatments} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N} \\
SS_{between \ subjects} = \Sigma \dfrac{P^2}{k} - \dfrac{G^2}{N} \\
SS_{error} = SS_{within}-SS_{between \ subjects}
$$

:::

::: {.column .smallest}

$$
df_{total} = N-1  \\
df_{within} = N-k \\
df_{between \ treatments} = k-1 \\
df_{between \ subjects} = n-1 \\
df_{error} = df_{within}-df_{between \ subjects}
$$

:::

:::: 

<!-- ## Calculations: $SS$ and $df$ {.smaller} -->

<!-- $$\begin{align} -->
<!-- SS_{total} &= \Sigma X^2 - \dfrac{G^2}{N} \\ -->
<!-- SS_{within} &= \Sigma SS_{within \ each \ treatment} \\ -->
<!-- SS_{between \ treatments} &= \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N} \\ -->
<!-- SS_{between \ subjects} &= \Sigma \dfrac{P^2}{k} - \dfrac{G^2}{N} \\ -->
<!-- SS_{error} &= SS_{within}-SS_{between \ subjects} -->
<!-- \end{align}$$ -->


<!-- $$\begin{align} -->
<!-- df_{total} &= N-1  \\ -->
<!-- df_{within} &= N-k \\ -->
<!-- df_{between \ treatments} &= k-1 \\ -->
<!-- df_{between \ subjects} &= n-1 \\ -->
<!-- df_{error} &= df_{within}-df_{between \ subjects} -->
<!-- \end{align}$$ -->

## Calculations: $MS$ and $F$ {.small}

- Step 2. $MS$ values

$$\begin{align}
MS_{between \ treatments} &= \dfrac{SS_{between \ treatments}}{df_{between \ treatments}} \\
\ \\
MS_{error} &= \dfrac{SS_{error}}{df_{error}}
\end{align}$$

- Step 3. $F$-ratio

::: {.center-element}
$F = \dfrac{MS_{between \ treatments}}{MS_{error}}$
:::

## Summary table

```{r}

s_tab <- data.frame(Source = c("Between treatments",
                               "Within treatments",
                               "&nbsp;&nbsp;&nbsp;Between subjects",
                               "&nbsp;&nbsp;&nbsp;Error",
                               "Total"),
                    SS = c("","","","",""),
                    df = c("","","","",""),
                    MS = c("","","","",""),
                    F = c("","","","",""))

knitr::kable(s_tab, format = 'html', escape = FALSE,
             col.names = c("Source", "\\(SS\\)", "\\(df\\)", "\\(MS\\)", "\\(F\\)"))
```

# Hypothesis test

## Step 1. State hypotheses

```{r data-setup}
options(digits = 2)

x <- c(9,11,13)
y <- c(3,5,4)
z <- c(5,6,7)

a2 <- bc1101tools::example_ANOVA(x, y, z)

N <- length(c(x, y, z))
n <- length(x)
k <- 3
G <- sum(c(x, y, z)) 
T1 <- sum(x)
T2 <- sum(y)
T3 <- sum(z)
P <- rowSums(data.frame(x, y, z))
sigma_x_sq <- sum(c(x,y,z)^2)
eta_sq <- a2$SS_between / a2$SS_total

```

```{r}
#| style: "font-size: 0.8em"

table2 <- knitr::kable(data.frame(person = c("A","B","C"), x, y, z), col.names = c("<br>Person", "🍌<br>Banana", "🍬<br>Candy", "😐<br>Control"), align = "c", format = "html", escape = FALSE) |> 
  kableExtra::add_header_above(c(" " = 1, "Manipulation" = 3)) |> 
  kableExtra::column_spec(1:4, width = "5em")

footer <- data.frame(quantity = c("\\(M\\)","\\(T\\)","\\(SS\\)"),
                    x = c(mean(x), sum(x), bc1101tools:::SS(x)),
                    y = c(mean(y), sum(y), bc1101tools:::SS(y)),
                    z = c(mean(z), sum(z), bc1101tools:::SS(z)),
                    P = "")

table3 <- knitr::kable(data.frame(person = c("A","B","C"), x, y, z, P), col.names = c("<br>Person", "🍌<br>Banana", "🍬<br>Candy", "😐<br>Control", "<br>P"), align = "c", format = "html", escape = FALSE) |> 
  kableExtra::add_header_above(c(" " = 1, "Manipulation" = 3, " " = 1)) |> 
  kableExtra::column_spec(1:4, width = "5em") |> 
  kableExtra::column_spec(5, width = "2em")

table_footer <- knitr::kable(footer, col.names = NULL, align = "c") |> 
  kableExtra::column_spec(1:4, width = "5em") |> 
  kableExtra::column_spec(5, width = "2em")


table2
```

- $H_0: \mu_1 = \mu_2 = \mu_3$
- $H_1$ : At least one population mean differs from another


## Step 2. Critical region {.small}

- Numerator: $df_{between \ treatments} = k-1$
- Denominator: $df_{error} = df_{within}-df_{between \ S's} = (N-k)-(n-1)$

```{r}
#| style: "font-size: 0.5em;"
options(digits = 3)

dfs <- setNames(1:10, 1:10)

aov_tab <- purrr::map_df(.x = dfs, ~qf(.95, df1 = 1:10, df2 = .x)) |> 
  t() |>
  as.data.frame() |>
  setNames(1:10) |>
  tibble::rownames_to_column(var = "df_denominator")


knitr::kable(aov_tab, format = "html", escape=FALSE, align = "c", digits = 3,
               col.names = c("\\(df_{denominator}\\)", 1:10)) |>
  kableExtra::add_header_above(c("\\(\\alpha = .05\\)", "\\(df_{numerator}\\)" = 10), escape=FALSE, align = "left")

options(digits = 2)
```


## 3. Calculate $F$-ratio


:::: {.columns}

::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em;"
# table3
# table_footer


table4 <- knitr::kable(data.frame(person = c("A","B","C", "\\(M\\)", "\\(T\\)", "\\(SS\\)"), 
                                  c(x, mean(x), sum(x), bc1101tools:::SS(x)),
                                  c(y, mean(y), sum(y), bc1101tools:::SS(y)),
                                  c(z, mean(z), sum(z), bc1101tools:::SS(z)),
                                  c(P, "", "", "")), 
                       col.names = c("<br>Person", "🍌<br>Banana", "🍬<br>Candy", "😐<br>Control", "<br>P"), 
                       align = "c", format = "html", escape = FALSE) |> 
  kableExtra::add_header_above(c(" " = 1, "Manipulation" = 3, " " = 1)) |> 
  kableExtra::column_spec(1:4, width = "5em") |> 
  kableExtra::column_spec(5, width = "2em") |> 
  kableExtra::row_spec(3, extra_css = "border-bottom: 3px solid black;")

table4
```

:::

::: {.column width="30%"}
$n = `r n`$  
$k = `r k`$  
$N = `r N`$  
$G = `r G`$  
$\Sigma X^2 = `r sigma_x_sq`$
:::

::::

::: {.smallest}
$$\begin{align}
df_{total} &= N-1 = `r N - 1` \\
df_{within} &= N-k  = `r N - k` \\
df_{between \ treatments} &= k-1  = `r k - 1` \\
df_{between \ subjects} &= n-1  = `r n - 1`\\
df_{error} &= df_{within}-df_{between \ subjects}  = `r (N - k) - (n - 1)`
\end{align}$$
:::



## 3. Calculate $F$-ratio


:::: {.columns}

::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em;"
table3
table_footer
```

:::

::: {.column width="30%"}
$n = `r n`$  
$k = `r k`$  
$N = `r N`$  
$G = `r G`$  
$\Sigma X^2 = `r sigma_x_sq`$
:::

::::

::: {.smallest}
:::: {.columns}

::: {.column width="50%"}

$SS_{total} = \Sigma X^2 - \dfrac{G^2}{N} = `r a2$SS_total`$

$SS_{within} = \Sigma SS_{within \ each \ treatment} = `r a2$SS_within`$

$SS_{between \ treatments} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N} = `r a2$SS_between`$
:::

::: {.column width="50%"}

$SS_{between \ subjects} = \Sigma \dfrac{P^2}{k} - \dfrac{G^2}{N} = `r a2$SS_between_subjects`$

$SS_{error} = \Sigma SS_{within}-SS_{between \ subjects} = `r a2$SS_error`$

:::

::::

:::


## 3. Calculate $F$-ratio


:::: {.columns}

::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em;"
table3
table_footer
```

:::

::: {.column width="30%"}
$n = `r n`$  
$k = `r k`$  
$N = `r N`$  
$G = `r G`$  
$\Sigma X^2 = `r sigma_x_sq`$
:::

::::

::: {.smaller .center-element}
$MS_{between \ treatments} = \dfrac{SS_{between \ treatments}}{df_{between \ treatments}} = `r a2$MS_between`$

$MS_{error} = \dfrac{SS_{error}}{df_{error}} = `r a2$MS_error`$
:::



## 3. Calculate $F$-ratio


:::: {.columns}

::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em;"
table3
table_footer
```

:::

::: {.column width="30%"}
$n = `r n`$  
$k = `r k`$  
$N = `r N`$  
$G = `r G`$  
$\Sigma X^2 = `r sigma_x_sq`$
:::

::::

::: {.center-element}
$F = \dfrac{MS_{between \ treatments}}{MS_{error}} = `r a2$f_related`$
:::

# Step 4. Decision

- Step 4a. Significance
  - $F > F_{critical}$?

- Step 4b. Effect size

$$\begin{align}
\eta^2_{partial} &= \dfrac{SS_{between \ treatments}}{SS_{total} - SS_{between \ subjects}} \\
&= \dfrac{SS_{between \ treatments}}{SS_{between \ treatments} + SS_{error}} \\
&= `r a2$eta_sq_partial`
\end{align}$$


# Post-hoc tests

```{r}
q <- qtukey(p = .95, nmeans = 3, df = a2$df_error)

q_df <- data.frame(df = 2:8)
for (i in 2:7) {
  q_df <- tibble::add_column(q_df, data.frame(x = qtukey(.95, i, q_df$df)))
}
```

- Step 4c: Post-hoc tests
  - Tukey’s Honestly Significant Difference
  - Minimum difference between pairs of treatment means so that $p < \alpha_{experimentwise}$

:::: {.columns}

::: {.column width="50%"}
$$\begin{align}
HSD &= q \sqrt{\dfrac{MS_{denominator}}{n}} \\
&=`r q` \sqrt{\dfrac{`r a2$MS_error`}{`r n`}} \\
&= `r q * sqrt(a2$MS_error/n)`
\end{align}$$
:::

::: {.column width="50%"}
```{r}
#| style: "font-size: 0.5em;"
options(digits = 3)
knitr::kable(q_df, col.names = c("\\(df\\)", 2:7)) |> 
  kableExtra::column_spec(1, bold = TRUE) |> 
  kableExtra::add_header_above(c(" " = 1, "Number of conditions" = 6))
options(digits = 2)
```

:::

::::


## Report results

> A single-factor, related-samples ANOVA revealed a significant difference in people's test scores when the test was preceded by consumption of a banana `r bc1101tools::descriptives(x)`, a candy bar `r bc1101tools::descriptives(y)`, and no snack `r bc1101tools::descriptives(z)`; $F(`r a2$df_between`,`r a2$df_error`) = `r a2$f_related`$, $p < .05$, $\eta^2 = `r a2$eta_sq_partial`$.
> Post-hoc tests using Tukey's HSD revealed that test scores were significantly better following banana consumption than following no snack or candy consumption; the candy did not differ significantly from the control condition.


# Learning checks


1. A researcher obtains an $F$-ratio with $df = 2, 12$ in an independent-samples study ANOVA
    - How many levels of the IV were there?
    - How many subjects participated in the study?<br><br>
2. A researcher obtains an $F$-ratio with $df = 2, 12$ in a repeated-measures study ANOVA
    - How many levels of the IV were there?
    - How many subjects participated in the study?
  

::: {.content-hidden}

# Data and figures

## Cover

```{r cover-ggplot}
#| eval: false
library(ggplot2)
x_seq <- seq(-5,5,.1)
means <- seq(-2, 2, 1)

dat <- purrr::map_df(means, ~data.frame(x = x_seq, y = dnorm(x_seq, mean = .), group = .))

cover <- ggplot(dat, aes(x = x, y = y, fill = group, group = group)) +
  geom_area(position = "dodge", alpha = .5) +
  scale_x_continuous(expand = c(0,0)) +
  theme_void() + 
  theme(legend.position = "none")
cover
bc1101tools::plot_save("17_related-samples-ANOVA.svg", subdir = "covers", plot = cover,
       width = 10, height = 5)

```

## OJS

:::

```{ojs}

w = 500
h = 550
r = 20
fill = d3.scaleOrdinal(d3.schemeSet3)

makeSample = function () {
    var arr = [];
    for (var i = 0; i < 10; i++) {
      arr.push(Math.floor(Math.random()*10))
    }
    return arr;
}

data = ["A","B","C"]
  
independentSamplesFigure = {

  const svg = d3.select("#independent-samples-container")
    .append("svg")
    .attr("viewBox", "0 0 " + w + " " + h)
    .attr("preserveAspectRatio", "xMinYMin meet")

  const labels = svg.append("g")
    .style("font-size", "20px")
    .style("font-weight", "bold")
    .attr("class", "invertable")
    
  labels.selectAll("text")
    .data(data).enter().append("text")
    .attr("x", (d,i) => w/4 + i * w/data.length).attr("y", 20)
    .text(d => "SAMPLE " + d).attr("text-anchor", "middle")

  const sampleDots = svg.append("g")
  
  const g = sampleDots.selectAll("g").data(data).enter().append("g").attr("transform", (d,i) => `translate(${w/4 + i * w/data.length}, 0)`).style("stroke", "black")
  
  g.each(function(d,i){
    d3.select(this)
    .selectAll("circle")
      .data(makeSample()).enter().append("circle")
        .attr("cy", (d,i) => 60 + i*r*2.5)
        .attr("r", r)
        .style("fill", d => fill(d))
  })
      
}


relatedSamplesFigure = {

  const svg = d3.selectAll("#related-samples-container")
    .append("svg")
    .attr("viewBox", "0 0 " + w + " " + h)
    .attr("preserveAspectRatio", "xMinYMin meet")

  const labels = svg.append("g")
    .style("font-size", "20px")
    .style("font-weight", "bold")
    .attr("class", "invertable")
    
  labels.selectAll("text")
    .data(data).enter().append("text")
    .attr("x", (d,i) => w/4 + i * w/data.length).attr("y", 20)
    .text(d => "SAMPLE " + d).attr("text-anchor", "middle")
    
  const lines = svg.append("g").attr("transform", `translate(0, 60)`).style("stroke", "black")
  .attr("class", "invertable")
    .selectAll("line").data([0,1,2,3,4,5,6,7,8,9]).enter()
    .append("line")
      .attr("x1", w/4 + r * 1.5)
      .attr("x2", w/4 + 2 * w/data.length - r*1.5)
      .attr("y1", (d,i) => i * r * 2.5)
      .attr("y2", (d,i) => i * r * 2.5)

  const sampleDots = svg.append("g")
  const sample = makeSample()
  
  const g = sampleDots.selectAll("g").data(data).enter().append("g").attr("transform", (d,i) => `translate(${w/4 + i * w/data.length}, 0)`).style("stroke", "black")
  
  g.each(function(d,i){
    d3.select(this)
    .selectAll("circle")
      .data(sample).enter().append("circle")
        .attr("cy", (d,i) => 60 + i*r*2.5)
        .attr("r", r)
        .style("fill", d => fill(d))
  })
      
}

matchedSamplesFigure = {

   function makeSample () {
    var arr = [];
    for (var i = 0; i < 10; i++) {
      arr.push(Math.floor(Math.random()*10))
    }
    return arr;
  }

  const svg = d3.selectAll("#matched-samples-container")
    .append("svg")
    .attr("viewBox", "0 0 " + w + " " + h)
    .attr("preserveAspectRatio", "xMinYMin meet")

  const labels = svg.append("g")
    .style("font-size", "20px")
    .style("font-weight", "bold")
    .attr("class", "invertable")
    
  labels.selectAll("text")
    .data(["A","B"]).enter().append("text")
    .attr("x", (d,i) => w/4 + i * w/2).attr("y", 20)
    .text(d => "SAMPLE " + d).attr("text-anchor", "middle")
    
  const lines = svg.append("g").attr("transform", `translate(${w/4 + r * 1.5}, 60)`)
    .style("stroke", "black").style("stroke-dasharray", [10,10]).attr("class", "invertable")
    .selectAll("line").data([0,1,2,3,4,5,6,7,8,9]).enter()
    .append("line")
      .attr("x2", w/2 - 2 * r * 1.5)
      .attr("y1", (d,i) => i * r * 2.5)
      .attr("y2", (d,i) => i * r * 2.5)

  const sampleDots = svg.append("g")
  const sample = makeSample()
  
  const g = sampleDots.selectAll("g").data([0,1]).enter().append("g").attr("transform", (d,i) => `translate(${w/4 + i * w/2}, 0)`).style("stroke", "black")
  
  g.each(function(d,i){
    d3.select(this)
    .selectAll("circle")
      .data(sample).enter().append("circle")
        .attr("cy", (d,i) => 60 + i*r*2.5)
        .attr("r", r)
        .style("fill", d => fill(d))
  })

}

```
