---
format: revealjs
---

# 18|FACTORIAL ANOVA

![](covers/15_ANOVA.svg)

```{r data}

banana_math <- c(7,9,8,9)
banana_vocab <- c(5,4,6,5)
snickers_math <- c(5,3,4,4)
snickers_vocab <- c(6,6,5,5)

dat <- data.frame(values = c(banana_math, banana_vocab, snickers_math, snickers_vocab),
                  IV1 = rep(c("Banana", "Candy Bar"), each = 8),
                  IV2 = rep(c("Math", "Reaction Time"), each = 4))

analysis <- aov(values ~ IV1 * IV2, data = dat)
# summary(analysis)

N <- nrow(dat)
n <- length(banana_math)
nA <- dplyr::filter(dat, IV1=="Banana") |>  nrow()
nB <- dplyr::filter(dat, IV2=="Math") |>  nrow()
k <- 4
kA <- 2
kB <- 2

dfbetween <- k-1
dfwithin <- N-k
dfA <- kA-1
dfB <- kB-1
dfAB <- dfbetween - dfA - dfB

```


# Overview

- [Logic]
- [Interpretation]
- [Calculations]
- [Learning checks]


# Logic

## Terminology

- Terminology
  - Factor
      - The variable that designates the groups being compared
      - Between-participants / within-participants
  - Levels
      - Individual conditions or values that make up a factor
  - Factorial design
      - A study that combines two (or more) factors, each with two (or more) levels
      - Can be manipulated between-participants or within-participants (or mixed-factorial design)
      - E.g. 2x2 between-participants ANOVA

## Logic

- Three hypotheses tested by three $F$-ratios
  1. Main effect of Factor A
  2. Main effect of Factor B
  3. Interaction between A and B

- Each tested with same basic $F$-ratio structure

$F = \dfrac{\textrm{variance between treatments}}{\textrm{variance expected with no treatment effect}}$


## Partitioning variance

:::: {#diagram-container style="position: relative; width: 80%; height: 300px; font-size: 0.6em; margin: 0 auto;"}

::: {.node .top .total}
Total variance
:::
::: {.node .between style="top: 50%; left: 8%"}
Variance between treatments   
$MS_{between \ treatments}$
:::

::: {.node .within style="top: 50%; right: 8%;"}
Variance within groups  
$MS_{within}$
:::
::: {.node style="top: 85%; left: 0; width: 18%;"}
Factor A  
$MS_{A}$
:::
::: {.node style="top: 85%; left: 20%; width: 18%"}
Factor B  
$MS_{B}$
:::
::: {.node style="top: 85%; left: 40%; width: 18%"}
Interaction  
$MS_{A*B}$
:::

::::


# Calculations {.smaller}

:::: {.columns}

::: {.column width="50%"}
$SS_{total} = \Sigma X^2 - \dfrac{G^2}{N}$  
$SS_{within} = \Sigma SS_{within \ each \ treatment}$  
$SS_{between} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N}$  
$SS_{A} = \Sigma \dfrac{T^2_{col}}{n_{col}} - \dfrac{G^2}{N}$  
$SS_{B} = \Sigma \dfrac{T^2_{row}}{n_{row}} - \dfrac{G^2}{N}$  
$SS_{A*B} = SS_{between}-SS_{A}-SS_{B}$
:::

::: {.column width="50%"}
$df_{total} = N-1$  
$df_{within} = N-k$  
$df_{between} = k-1$  
$df_{A} = k_A-1$  
$df_{B} = k_B-1$  
$df_{A*B} = df_{between}-df_A-df_B$
:::

::::

## Summary table

# Hypothesis test

## Step 1. Hypotheses

```{r}
# table
```

- Main effect of Snack Type
    - $H_0$: $\mu_{banana} = \mu_{candy}$
    - $H_1$: $\mu_{banana} \ne \mu_{candy}$
- Main effect of Test Type
    - $H_0$: $\mu_{math} = \mu_{RT}$
    - $H_1$: $\mu_{math} \ne \mu_{RT}$
- Snack * Test interaction
    - $H_0$: No interaction
    - $H_1$: There is an interaction


## Step 2. Critical region

- Step 2. Critical region
  - Numerators: $df_{A} = k_A-1 = 1 \\ df_{B} = k_B-1 = 1 \\ df_{A*B} = k-1 - df_A - df_B = 10 = 1$
  - Denominator: $df_{within} = N-k  = `r N - k`$


# Interpretation

## Interpreting interaction graphs

![](media/anova-2x2-generic-1.svg){.invertable}

- Slope indicates main effect of factor on x-axis
- Distance between lines indicates main effect of other factor
- Parallel lines indicate no interaction


## Interpreting interaction graphs

![](media/anova-2x2-generic-2.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-3.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-4.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-5.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-6.svg){.invertable}

## Your turn

:::: {.columns}

::: {.column width="50%"}
- Come up with your own example
  - 2 IVs with 2 levels each
  - 1 DV (the thing you measure, e.g. test performance)
  - Sketch graph of expected results
:::

::: {.column width="50%"}
![](media/anova-2x2-generic-blank.svg){.invertable}
:::

::::





# Learning checks

- True or False?
  - Two separate single-factor ANOVAs provide exactly the same information that is obtained from a two-factor analysis of variance
  - A disadvantage of combining 2 factors in an experiment is that you cannot determine how either factor would affect participants’ scores if it were examined in an experiment by itself
  - If a two-factor analysis of variance produces a statistically significant interaction, then you can conclude that either/both main effects for factor A or factor B are also significant




::: {.content-hidden}

# Data and figures

## Cover

## Generic 2x2 graphs

```{r}
#| eval: false
library(bc1101tools)

width <- 5
height <- 4

plot_2x2_generic(1, 2, 1, 2)
plot_save("anova-2x2-generic-1.svg", width = width, height = height)

plot_2x2_generic(.5, 1.5, 1.5, 2.5)
plot_save("anova-2x2-generic-2.svg", width = width, height = height)

plot_2x2_generic(2.5, 1.5, 1.5, 0.5)
plot_save("anova-2x2-generic-3.svg", width = width, height = height)

plot_2x2_generic(1.6, 1.4, 1.6, 1.4)
plot_save("anova-2x2-generic-4.svg", width = width, height = height)

plot_2x2_generic(1, 2, 2, 1)
plot_save("anova-2x2-generic-5.svg", width = width, height = height)

plot_2x2_generic(1.5, 2, 1.5, 1)
plot_save("anova-2x2-generic-6.svg", width = width, height = height)

plot_2x2_generic(0, 0, 0, 0)
plot_save("anova-2x2-generic-blank.svg", width = width, height = height)
```

:::
