---
format: revealjs
---

# 18|FACTORIAL ANOVA

![](covers/15_ANOVA.svg)

```{r data}
options(digits = 2)

banana_math <- c(7,9,8,9)
banana_rt <- c(5,4,6,5)
candy_math <- c(5,3,4,4)
candy_rt <- c(6,6,5,5)

df <- data.frame(A = rep(c(1,2), each = 8),
                 B = rep(c(1,2), each = 4),
                 DV = c(banana_math, banana_rt, candy_math, candy_rt))

a <- bc1101tools::example_ANOVA_factorial(df)

table_scores <- bc1101tools::anova_2x2_table(
  factor_a_name = "Snack", 
  factor_a_levels = c("Banana", "Candy"),
  factor_b_name = "Test",
  factor_b_levels = c("Math", "Reaction time"),
  cell_values = c(paste(banana_math, collapse = ", "),  
                  paste(banana_rt, collapse = ", "), 
                  paste(candy_math, collapse = ", "), 
                  paste(candy_rt, collapse = ", "))
  
)

cell_values <- c(
  glue::glue("\\(T = {a$T1}\\)<br>\\(SS = {a$SS1}\\)"),
  glue::glue("\\(T = {a$T2}\\)<br>\\(SS = {a$SS2}\\)"),
  glue::glue("\\(T = {a$T3}\\)<br>\\(SS = {a$SS3}\\)"),
  glue::glue("\\(T = {a$T4}\\)<br>\\(SS = {a$SS4}\\)")
)

table_summary <- bc1101tools::anova_2x2_table(
  factor_a_name = "Snack", 
  factor_a_levels = c("Banana", "Candy"),
  factor_b_name = "Test",
  factor_b_levels = c("Math", "Reaction time"),
  cell_values = cell_values
)

```


# Overview

- [Logic]
- [Calculations]
- [Hypothesis test]
- [Interpretation]
- [Learning checks]


# Logic

## Terminology {.small}

- Factor
  - The variable that designates the groups being compared
  - Between-participants / within-participants
- Levels
  - Individual conditions or values that make up a factor
- Factorial design
  - A study that combines two (or more) factors, each with two (or more) levels
  - Can be manipulated between-participants or within-participants (or mixed-factorial design)
  - E.g. 2x2 between-participants ANOVA

## Logic

```{r}
bc1101tools::anova_2x2_table()
```

  
- Three hypotheses tested by three $F$-ratios
  1. Main effect of Factor A
  2. Main effect of Factor B
  3. Interaction between A and B

- Each tested with same basic $F$-ratio structure

$F = \dfrac{\textrm{variance between treatments}}{\textrm{variance expected with no treatment effect}}$


## Partitioning variance

:::: {#diagram-container style="position: relative; width: 80%; height: 300px; font-size: 0.6em; margin: 0 auto;"}

::: {.node .top .total}
Total variance
:::
::: {.node .between style="top: 50%; left: 8%"}
Variance between treatments   
$MS_{between \ treatments}$
:::

::: {.node .within style="top: 50%; right: 8%;"}
Variance within groups  
$MS_{within}$
:::
::: {.node style="top: 85%; left: 0; width: 18%;"}
Factor A  
$MS_{A}$
:::
::: {.node style="top: 85%; left: 20%; width: 18%"}
Factor B  
$MS_{B}$
:::
::: {.node style="top: 85%; left: 40%; width: 18%"}
Interaction  
$MS_{A*B}$
:::

::::



# Calculations {.smaller}

## Example data

```{r}
table_scores
```



## Equations

:::: {.columns}

::: {.column width="50%"}
$SS_{total} = \Sigma X^2 - \dfrac{G^2}{N}$  
$SS_{within} = \Sigma SS_{within \ each \ treatment}$  
$SS_{between} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N}$  
$SS_{A} = \Sigma \dfrac{T^2_{col}}{n_{col}} - \dfrac{G^2}{N}$  
$SS_{B} = \Sigma \dfrac{T^2_{row}}{n_{row}} - \dfrac{G^2}{N}$  
$SS_{A*B} = SS_{between}-SS_{A}-SS_{B}$
:::

::: {.column width="50%"}
$df_{total} = N-1$  
$df_{within} = N-k$  
$df_{between} = k-1$  
$df_{A} = k_A-1$  
$df_{B} = k_B-1$  
$df_{A*B} = df_{between}-df_A-df_B$
:::

::::

## Summary table

# Hypothesis test

## Step 1. Hypotheses

```{r}
table_scores
```

- Main effect of Snack Type
    - $H_0$: $\mu_{banana} = \mu_{candy}$
    - $H_1$: $\mu_{banana} \ne \mu_{candy}$
- Main effect of Test Type
    - $H_0$: $\mu_{math} = \mu_{RT}$
    - $H_1$: $\mu_{math} \ne \mu_{RT}$
- Snack * Test interaction
    - $H_0$: No interaction
    - $H_1$: There is an interaction


## Step 2. Critical region

- Step 2. Critical region
  - Numerators: $df_{A} = k_A-1 = 1 \\ df_{B} = k_B-1 = 1 \\ df_{A*B} = k-1 - df_A - df_B = 1$
  - Denominator: $df_{within} = N-k  = `r a$N - a$k`$


## Step 3. Calculations

:::: {.columns}
::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```
:::

::: {.column style="width: 30%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

$$\begin{align}
df_{total} &= N-1 = `r a$N - 1` \\
df_{within} &= N-k  = `r a$N - a$k` \\
df_{between} &= k-1  = `r a$k - 1`\\
df_{A} &= k_A-1  = `r a$df_a`\\
df_{B} &= k_B-1  = `r a$df_b`\\
df_{A*B} &= df_{between} - df_A - df_B  = `r a$df_ab`\\
\end{align}$$


## Step 3. Calculations

:::: {.columns}
::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```
:::

::: {.column style="width: 30%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

$$\begin{align}
SS_{total} &= \Sigma X^2 - \dfrac{G^2}{N} = `r a$ss_total`\\
SS_{within} &= \Sigma SS_{within \ each \ treatment} = `r a$ss_within`\\
SS_{between} &= \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N} = `r a$ss_between` \\
\end{align}$$


## Step 3. Calculations

:::: {.columns}
::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```
:::

::: {.column style="width: 30%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

$$\begin{align}
SS_{A} &= \Sigma \dfrac{T^2_{col}}{n_{col}} - \dfrac{G^2}{N} = `r a$ss_a`\\
SS_{B} &= \Sigma \dfrac{T^2_{row}}{n_{row}} - \dfrac{G^2}{N} = `r a$ss_b`\\
SS_{A*B} &= \Sigma SS_{between}-SS_A - SS_B = `r a$ss_ab`
\end{align}$$


## Step 3. Calculations

:::: {.columns}
::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```
:::

::: {.column style="width: 30%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

::: {.center-element}
$MS_{A} = \dfrac{SS_{A}}{df_{A}} = `r a$ms_a` \ \ \ \ \ \ \ \ MS_{B} = \dfrac{SS_{B}}{df_{B}} = `r a$ms_b`$

$MS_{A*B} = \dfrac{SS_{A*B}}{df_{A*B}} = `r a$ms_ab`$

$MS_{within} = \dfrac{SS_{within}}{df_{within}} = `r a$ms_within`$
:::

## Step 3. Calculations

:::: {.columns}
::: {.column width="70%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```
:::

::: {.column style="width: 30%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

::: {.center-element}
$F_A = \dfrac{MS_{A}}{MS_{within}} = `r a$f_a`$

$F_B = \dfrac{MS_{B}}{MS_{within}} = `r a$f_b`$

$F_{A*B} = \dfrac{MS_{A*B}}{MS_{within}} = `r a$f_ab`$
:::


# Interpretation

## Interpreting interaction graphs

![](media/anova-2x2-generic-1.svg){.invertable}

- Slope indicates main effect of factor on x-axis
- Distance between lines indicates main effect of other factor
- Parallel lines indicate no interaction


## Interpreting interaction graphs

![](media/anova-2x2-generic-2.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-3.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-4.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-5.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-6.svg){.invertable}

## Your turn

:::: {.columns}

::: {.column width="50%"}
- Come up with your own example
  - 2 IVs with 2 levels each
  - 1 DV (the thing you measure, e.g. test performance)
  - Sketch graph of expected results
:::

::: {.column width="50%"}
![](media/anova-2x2-generic-blank.svg){.invertable}
:::

::::





# Learning checks {.small}

- True or False?
  - Two separate single-factor ANOVAs provide exactly the same information that is obtained from a two-factor analysis of variance
  - A disadvantage of combining 2 factors in an experiment is that you cannot determine how either factor would affect participants’ scores if it were examined in an experiment by itself
  - If a two-factor analysis of variance produces a statistically significant interaction, then you can conclude that either/both main effects for factor A or factor B are also significant




::: {.content-hidden}

# Data and figures

## Cover

## Generic 2x2 graphs

```{r}
#| eval: false
library(bc1101tools)

width <- 5
height <- 4

plot_2x2_generic(1, 2, 1, 2)
plot_save("anova-2x2-generic-1.svg", width = width, height = height)

plot_2x2_generic(.5, 1.5, 1.5, 2.5)
plot_save("anova-2x2-generic-2.svg", width = width, height = height)

plot_2x2_generic(2.5, 1.5, 1.5, 0.5)
plot_save("anova-2x2-generic-3.svg", width = width, height = height)

plot_2x2_generic(1.6, 1.4, 1.6, 1.4)
plot_save("anova-2x2-generic-4.svg", width = width, height = height)

plot_2x2_generic(1, 2, 2, 1)
plot_save("anova-2x2-generic-5.svg", width = width, height = height)

plot_2x2_generic(1.5, 2, 1.5, 1)
plot_save("anova-2x2-generic-6.svg", width = width, height = height)

plot_2x2_generic(0, 0, 0, 0)
plot_save("anova-2x2-generic-blank.svg", width = width, height = height)
```

:::
