---
format: revealjs
---

# 18|FACTORIAL ANOVA

![](covers/18_factorial-ANOVA.svg)

```{r data}
options(digits = 2)

banana_math <- c(7,9,8,9)
banana_rt <- c(5,4,6,5)
candy_math <- c(5,3,4,4)
candy_rt <- c(6,6,5,5)

df <- data.frame(A = rep(c(1,2), each = 8),
                 B = rep(c(1,2), each = 4),
                 DV = c(banana_math, banana_rt, candy_math, candy_rt))

a <- bc1101tools::example_ANOVA_factorial(df)

table_scores <- bc1101tools::anova_2x2_table(
  factor_a_name = "Snack", 
  factor_a_levels = c("Banana", "Candy"),
  factor_b_name = "Test",
  factor_b_levels = c("Math", "Reaction time"),
  cell_values = c(paste(banana_math, collapse = ", "),  
                  paste(banana_rt, collapse = ", "), 
                  paste(candy_math, collapse = ", "), 
                  paste(candy_rt, collapse = ", "))
  
)

cell_values <- c(
  glue::glue("\\(T = {a$T1}\\)<br>\\(SS = {a$SS1}\\)"),
  glue::glue("\\(T = {a$T2}\\)<br>\\(SS = {a$SS2}\\)"),
  glue::glue("\\(T = {a$T3}\\)<br>\\(SS = {a$SS3}\\)"),
  glue::glue("\\(T = {a$T4}\\)<br>\\(SS = {a$SS4}\\)")
)

table_summary <- bc1101tools::anova_2x2_table(
  factor_a_name = "Snack", 
  factor_a_levels = c("Banana", "Candy"),
  factor_b_name = "Test",
  factor_b_levels = c("Math", "Reaction time"),
  cell_values = cell_values
)

```


# Overview

- [Logic]
- [Calculations]
- [Hypothesis test]
- [Interpretation]
- [Learning checks]


# Logic

## Terminology {.small}

- Factor
  - The variable that designates the groups being compared
  - Between-participants / within-participants
- Levels
  - Individual conditions or values that make up a factor
- Factorial design
  - A study that combines two (or more) factors, each with two (or more) levels
  - Can be manipulated between-participants or within-participants (or mixed-factorial design)
  - E.g. 2x2 between-participants ANOVA

## Logic

```{r}
bc1101tools::anova_2x2_table()
```

  
- Three hypotheses tested by three $F$-ratios
  1. Main effect of Factor A
  2. Main effect of Factor B
  3. Interaction between A and B


## Logic

```{r}
bc1101tools::anova_2x2_table()
```

  
- Three hypotheses tested by three $F$-ratios
  - Each tested with same basic $F$-ratio structure

::: {.center-element}
$F = \dfrac{\textrm{variance between treatments}}{\textrm{variance expected with no treatment effect}}$
:::


# Calculations {.smaller}

## Example data

```{r}
table_scores
```


## Partitioning variance

<style>

  .node.level3:before {
    content: "";
    position: absolute;
    width: 3em;
    height: 0;
    border: 3px dashed var(--text-color);
    <!-- z-index: -10; -->
  }
  .node.msa:before {
    transform: translate(30%, -2.2em) rotate(-45deg) scale(0.5);
  }
  .node.msb:before {
    transform: translate(0%, -2.1em) rotate(90deg) scale(0.5);
    width: 2.5em;
    z-index: -2;
  }
  .node.msab:before {
    transform: translate(-30%, -2.2em) rotate(45deg) scale(0.5);
  }

</style>

:::: {#diagram-container style="position: relative; width: 80%; height: 300px; font-size: 0.6em; margin: 0 auto;"}

::: {.node .top .total}
Total variance
:::
::: {.node .between style="top: 50%; left: 8%;"}
Variance between treatments   
$MS_{between \ treatments}$
:::

::: {.node .within style="top: 50%; right: 8%;"}
Variance within groups  
$MS_{within}$
:::

::: {.fragment}

::: {.node .level3 .msa style="top: 85%; left: 0; width: 18%;"}
Factor A  
$MS_{A}$
:::
::: {.node .level3 .msb style="top: 85%; left: 20%; width: 18%"}
Factor B  
$MS_{B}$
:::
::: {.node .level3 .msab style="top: 85%; left: 40%; width: 18%"}
Interaction  
$MS_{A*B}$
:::
:::

::::


## Partitioning variance

<style>
  .xx-total {
    display: table;
    margin: 0 auto;
    margin-bottom: 2em;
  }
  .xx-total:before {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(135deg) scale(0.5);
  }
  .xx-total:after {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(45deg) scale(0.5);
  }
  .xx-between {
    float: left;
    margin-left: 20%;
    margin-right: -4em;
    margin-bottom: 2em;
  }
  .xx-within {
    float: right;
    margin-right: 20%;
    <!-- margin-top: -2.5em; -->
    <!-- margin-bottom: 4em; -->
  }
  .xx-between:before {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(135deg) scale(0.5);
  }
  .xx-between:after {
    content: "";
    position: absolute;
    width: 4em;
    height: 0;
    border: 3px dashed var(--text-color);
    transform: translate(-50%, 2.3em) rotate(45deg) scale(0.5);
  }

  .ss-level3 {
    white-space: nowrap;
    position: relative;
    transform: translate(0, 3em);
  }
  .xx-b:before {
    content: "";
    position: absolute;
    width: 1.3em;
    height: 0;
    border: 1.5px dashed var(--text-color);
    transform: translate(-0.2em, -0.7em) rotate(90deg);
  }

</style>

:::: {.columns style="font-size: 0.7em;"}

::: {.column width="50%"}

[$SS_{total}$]{.xx-total}

[$SS_{between}$]{.xx-between} [$SS_{within}$]{.xx-within}

<!-- [$SS_A$ &nbsp;&nbsp;&nbsp; $SS_B$ &nbsp;&nbsp;&nbsp; $SS_{A*B}$]{.ss-diagram .xx-ab .level3} -->

::: {.ss-level3}
[$SS_A$]{style="margin-left: -1em;margin-right: 1em;"} [$SS_B$]{.xx-b} [$SS_{A*B}$]{style="margin-left: 1em;"} 
:::

:::

::: {.column width="50%"}

[$df_{total}$]{.xx-total}

[$df_{between}$]{.xx-between} [$df_{within}$]{.xx-within}

::: {.ss-level3}
[$df_A$]{style="margin-left: -1em;margin-right: 1.2em;"} [$df_B$]{.xx-b} [$df_{A*B}$]{style="margin-left: 1em;"} 
:::

:::

::::

:::: {.columns .center-element}

::: {.column .smallest style="margin-top: 1em;"}

$SS_{total} = \Sigma X^2 - \dfrac{G^2}{N}$  
$SS_{within} = \Sigma SS_{within \ each \ treatment}$  
$SS_{between} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N}$  
$SS_{A} = \Sigma \dfrac{T^2_{col}}{n_{col}} - \dfrac{G^2}{N}$  
$SS_{B} = \Sigma \dfrac{T^2_{row}}{n_{row}} - \dfrac{G^2}{N}$  
$SS_{A*B} = SS_{between}-SS_{A}-SS_{B}$

:::

::: {.column .smallest style="margin-top: 3em;"}

$df_{total} = N-1$  
$df_{within} = N-k$  
$df_{between} = k-1$  
$df_{A} = k_A-1$  
$df_{B} = k_B-1$  
$df_{A*B} = df_{between}-df_A-df_B$

:::

:::: 


<!-- ## Equations -->

<!-- :::: {.columns} -->

<!-- ::: {.column width="50%"} -->
<!-- $SS_{total} = \Sigma X^2 - \dfrac{G^2}{N}$   -->
<!-- $SS_{within} = \Sigma SS_{within \ each \ treatment}$   -->
<!-- $SS_{between} = \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N}$   -->
<!-- $SS_{A} = \Sigma \dfrac{T^2_{col}}{n_{col}} - \dfrac{G^2}{N}$   -->
<!-- $SS_{B} = \Sigma \dfrac{T^2_{row}}{n_{row}} - \dfrac{G^2}{N}$   -->
<!-- $SS_{A*B} = SS_{between}-SS_{A}-SS_{B}$ -->
<!-- ::: -->

<!-- ::: {.column width="50%"} -->
<!-- $df_{total} = N-1$   -->
<!-- $df_{within} = N-k$   -->
<!-- $df_{between} = k-1$   -->
<!-- $df_{A} = k_A-1$   -->
<!-- $df_{B} = k_B-1$   -->
<!-- $df_{A*B} = df_{between}-df_A-df_B$ -->
<!-- ::: -->

<!-- :::: -->

## Summary table

```{r}
#| style: "width: 100%;"

s_tab <- data.frame(Source = c("Between treatments",
                               "&nbsp;&nbsp;&nbsp;Between subjects",
                               "&nbsp;&nbsp;&nbsp;Between subjects",
                               "&nbsp;&nbsp;&nbsp;Error",
                               "Within treatments",
                               "Total"),
                    SS = c("","","","","",""),
                    df = c("","","","","",""),
                    MS = c("","","","","",""),
                    F = c("","","","","",""))

knitr::kable(s_tab, format = 'html', escape = FALSE,
             col.names = c("Source", "\\(SS\\)", "\\(df\\)", "\\(MS\\)", "\\(F\\)")) |> 
  kableExtra::column_spec(1, width = "10em") |> 
  kableExtra::column_spec(2:5, width = "2em")
```

# Hypothesis test

## Step 1. Hypotheses

```{r}
#| style: "font-size: 0.7em;"
table_scores
```

::: {style="font-size: 0.6em;"}
- Main effect of Snack Type
    - $H_0$: $\mu_{banana} = \mu_{candy}$
    - $H_1$: $\mu_{banana} \ne \mu_{candy}$
- Main effect of Test Type
    - $H_0$: $\mu_{math} = \mu_{RT}$
    - $H_1$: $\mu_{math} \ne \mu_{RT}$
- Snack * Test interaction
    - $H_0$: No interaction
    - $H_1$: There is an interaction
:::

## Step 2. Critical region {.smaller}

- Step 2. Critical region
  - Numerators: $df_{A} = k_A-1 = 1 \\ df_{B} = k_B-1 = 1 \\ df_{A*B} = k-1 - df_A - df_B = 1$
  - Denominator: $df_{within} = N-k  = `r a$N - a$k`$

```{r}
#| style: "font-size: 0.6em;"
options(digits = 3)

dfs <- setNames(5:15, 5:15)

aov_tab <- purrr::map_df(.x = dfs, ~qf(.95, df1 = 1:10, df2 = .x)) |> 
  t() |>
  as.data.frame() |>
  setNames(1:10) |>
  tibble::rownames_to_column(var = "df_denominator")


knitr::kable(aov_tab, format = "html", escape=FALSE, align = "c", digits = 3,
               col.names = c("\\(df_{denominator}\\)", 1:10)) |>
  kableExtra::add_header_above(c("\\(\\alpha = .05\\)", "\\(df_{numerator}\\)" = 10), escape=FALSE, align = "left")

options(digits = 2)
```

## Step 3. Calculations

:::: {.columns}
::: {.column width="60%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```

::: {style="margin-left: 7em; font-size: 0.8em;"}
$T_{col}$  [$53$]{style="margin-left: 2em;"} [$38$]{style="margin-left: 4em;"}
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}

::: {style="margin-top: 0.8em; padding-top: 1em; height: 2em; display: flex; flex-direction: column; justify-content: space-between;"}
$T_{row}$

$49$

$42$
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

::: {style="font-size: 0.8em;"}
$$\begin{align}
df_{total} &= N-1 = `r a$N - 1` \\
df_{within} &= N-k  = `r a$N - a$k` \\
df_{between} &= k-1  = `r a$k - 1`\\
df_{A} &= k_A-1  = `r a$df_a`\\
df_{B} &= k_B-1  = `r a$df_b`\\
df_{A*B} &= df_{between} - df_A - df_B  = `r a$df_ab`\\
\end{align}$$
:::

## Step 3. Calculations

:::: {.columns}
::: {.column width="60%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```

::: {style="margin-left: 7em; font-size: 0.8em;"}
$T_{col}$  [$53$]{style="margin-left: 2em;"} [$38$]{style="margin-left: 4em;"}
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}

::: {style="margin-top: 0.8em; padding-top: 1em; height: 2em; display: flex; flex-direction: column; justify-content: space-between;"}
$T_{row}$

$49$

$42$
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

::: {style="font-size: 0.8em;"}
$$\begin{align}
SS_{total} &= \Sigma X^2 - \dfrac{G^2}{N} = `r a$ss_total`\\
SS_{within} &= \Sigma SS_{within \ each \ treatment} = `r a$ss_within`\\
SS_{between} &= \Sigma \dfrac{T^2}{n} - \dfrac{G^2}{N} = `r a$ss_between` \\
\end{align}$$
:::

## Step 3. Calculations

:::: {.columns}
::: {.column width="60%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```

::: {style="margin-left: 7em; font-size: 0.8em;"}
$T_{col}$  [$53$]{style="margin-left: 2em;"} [$38$]{style="margin-left: 4em;"}
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}

::: {style="margin-top: 0.8em; padding-top: 1em; height: 2em; display: flex; flex-direction: column; justify-content: space-between;"}
$T_{row}$

$49$

$42$
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

::: {style="font-size: 0.8em;"}
$$\begin{align}
SS_{A} &= \Sigma \dfrac{T^2_{col}}{n_{col}} - \dfrac{G^2}{N} = `r a$ss_a`\\
SS_{B} &= \Sigma \dfrac{T^2_{row}}{n_{row}} - \dfrac{G^2}{N} = `r a$ss_b`\\
SS_{A*B} &= \Sigma SS_{between}-SS_A - SS_B = `r a$ss_ab`
\end{align}$$
:::

## Step 3. Calculations

:::: {.columns}
::: {.column width="60%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```

::: {style="margin-left: 7em; font-size: 0.8em;"}
$T_{col}$  [$53$]{style="margin-left: 2em;"} [$38$]{style="margin-left: 4em;"}
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}

::: {style="margin-top: 0.8em; padding-top: 1em; height: 2em; display: flex; flex-direction: column; justify-content: space-between;"}
$T_{row}$

$49$

$42$
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

::: {.center-element style="font-size: 0.8em;"}
$MS_{A} = \dfrac{SS_{A}}{df_{A}} = `r a$ms_a` \ \ \ \ \ \ \ \ MS_{B} = \dfrac{SS_{B}}{df_{B}} = `r a$ms_b`$

$MS_{A*B} = \dfrac{SS_{A*B}}{df_{A*B}} = `r a$ms_ab`$

$MS_{within} = \dfrac{SS_{within}}{df_{within}} = `r a$ms_within`$
:::

## Step 3. Calculations

:::: {.columns}
::: {.column width="60%"}
```{r}
#| style: "font-size: 0.6em; margin-top: 1em;"

table_summary
```

::: {style="margin-left: 7em; font-size: 0.8em;"}
$T_{col}$  [$53$]{style="margin-left: 2em;"} [$38$]{style="margin-left: 4em;"}
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}

::: {style="margin-top: 0.8em; padding-top: 1em; height: 2em; display: flex; flex-direction: column; justify-content: space-between;"}
$T_{row}$

$49$

$42$
:::
:::

::: {.column style="width: 20%; font-size: 0.8em;"}
$N = `r a$N`$  
$n = `r a$n`$  
$k = `r a$k`$  
$k_A = `r a$k_a`$  
$k_B = `r a$k_b`$  
$G = `r a$G`$  
$\Sigma X^2 = `r a$sigma_x_sq`$
:::
::::

::: {.center-element style="font-size: 0.8em;"}
$F_A = \dfrac{MS_{A}}{MS_{within}} = `r a$f_a`$

$F_B = \dfrac{MS_{B}}{MS_{within}} = `r a$f_b`$

$F_{A*B} = \dfrac{MS_{A*B}}{MS_{within}} = `r a$f_ab`$
:::


## Step 4a. Decision

- For each $F$-ratio, reject or fail to reject $H_0$
  - Compare calculated $F$ to corresponding $F_{critical}$

![](media/2x2-snack-test-example-plot.svg){.invertable}

## Step 4b. Effect size

```{r}
eta_a <-  a$ss_a / (a$ss_total - a$ss_b - a$ss_ab)
eta_b <-  a$ss_b / (a$ss_total - a$ss_a - a$ss_ab)
eta_ab <- a$ss_ab / (a$ss_total - a$ss_a - a$ss_b)
```

- $\eta^2_{partial}$
  - Percentage of variability not explained by other factors
  - Separate effect size for each $F$-ratio  

::: {.center-element}
$\eta^2_A = \dfrac{SS_A}{SS_{total}-SS_{B} - SS_{A*B}} = `r eta_a`$

$\eta^2_B = \dfrac{SS_B}{SS_{total}-SS_{A} - SS_{A*B}} = `r eta_b`$

$\eta^2_{A*B} = \dfrac{SS_{A*B}}{SS_{total}-SS_{A} - SS_{B}} = `r eta_ab`$
:::


## Step 5. Report results

- Descriptives (usually in a table or graph)
- Results of hypothesis test for all three tests

:::: {.columns}

::: {.column style="font-size: 0.5em;"}
> To examine the influence of snack and test type on performance, a 2-factor ANOVA was conducted with test scores as the dependent variable and Snack Type and Test Type as between-participants independent variables. There was no significant main effect of Test Type $(F (1, 12) = 4.74$, $p > .05)$. There was, however, significant main effect of Snack Type $(F (1, 12) = 21.77$, $p < .05$, $\eta^2 = .64)$; overall, performance was superior in the Banana condition. Moreover, there was a significant interaction between Snack Type and Test Type $(F (1, 12) = 34.94$, $p < .05$, $\eta^2 = .74)$; performance on the math test was affected by snack type to a greater extent than performance on the reaction time test. The pattern of means and standard deviations is shown in Table 1. Trends are illustrated in Figure 1.
:::

::: {.column}
![Figure 1. Test performance by snack and test type](media/2x2-snack-test-example-plot.svg){.invertable}
:::

::::

# Interpretation

## Interpreting interaction graphs

![](media/anova-2x2-generic-1.svg){.invertable}

- Slope indicates main effect of factor on x-axis
- Distance between lines indicates main effect of other factor
- Parallel lines indicate no interaction


## Interpreting interaction graphs

![](media/anova-2x2-generic-2.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-3.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-4.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-5.svg){.invertable}

## Interpreting interaction graphs

![](media/anova-2x2-generic-6.svg){.invertable}

## Your turn

:::: {.columns}

::: {.column width="50%"}
- Come up with your own example
  - 2 IVs with 2 levels each
  - 1 DV (the thing you measure, e.g. test performance)
  - Sketch graph of expected results
:::

::: {.column width="50%"}
![](media/anova-2x2-generic-blank.svg){.invertable}
:::

::::





# Learning checks {.small}

- True or False?
  - Two separate single-factor ANOVAs provide exactly the same information that is obtained from a two-factor analysis of variance
  - A disadvantage of combining 2 factors in an experiment is that you cannot determine how either factor would affect participants’ scores if it were examined in an experiment by itself
  - If a two-factor analysis of variance produces a statistically significant interaction, then you can conclude that either/both main effects for factor A or factor B are also significant




::: {.content-hidden}

# Data and figures

## Cover

```{r cover-ggplot}
#| eval: false
library(ggplot2)
x_seq <- seq(-5,5,.1)
means <- seq(-2, 2, length.out = 4)

dat <- purrr::map_df(means, ~data.frame(x = x_seq, y = dnorm(x_seq, mean = .), group = .))

cover <- ggplot(dat, aes(x = x, y = y, fill = factor(group), group = group)) +
  geom_area(position = "dodge", alpha = .5) +
  scale_x_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("dodgerblue3", "red", "dodgerblue3", "red")) +
  theme_void() + 
  theme(legend.position = "none")

cover

bc1101tools::plot_save("18_factorial-ANOVA.svg", 
                       subdir = "covers", 
                       plot = cover,
                       width = 10, height = 5)

```

## 2x2 for snack * test type

```{r}
#| eval: false

library(ggplot2)
library(dplyr)

df |> 
  summarize(mean = mean(DV), .by = c(A, B)) |> 
  ggplot(aes(x = factor(A), y = mean, group = factor(B), linetype = factor(B))) +
  geom_line() +
  geom_point() +
  scale_x_discrete(labels = c("Banana", "Candy")) +
  scale_linetype(labels = c("Math", "Reaction time")) +
  labs(x = "Snack type", y = "Test score", linetype = "Test type") +
  bc1101tools::theme_bc1101() +
  theme(legend.key = element_rect(fill = NA),
        legend.background = element_rect(fill = NA),
        legend.box.background = element_rect(fill = NA, color = NA))

bc1101tools::plot_save("2x2-snack-test-example-plot.svg")
```


## Generic 2x2 graphs

```{r}
#| eval: false
library(bc1101tools)

width <- 5
height <- 4

plot_2x2_generic(1, 2, 1, 2)
plot_save("anova-2x2-generic-1.svg", width = width, height = height)

plot_2x2_generic(.5, 1.5, 1.5, 2.5)
plot_save("anova-2x2-generic-2.svg", width = width, height = height)

plot_2x2_generic(2.5, 1.5, 1.5, 0.5)
plot_save("anova-2x2-generic-3.svg", width = width, height = height)

plot_2x2_generic(1.6, 1.4, 1.6, 1.4)
plot_save("anova-2x2-generic-4.svg", width = width, height = height)

plot_2x2_generic(1, 2, 2, 1)
plot_save("anova-2x2-generic-5.svg", width = width, height = height)

plot_2x2_generic(1.5, 2, 1.5, 1)
plot_save("anova-2x2-generic-6.svg", width = width, height = height)

plot_2x2_generic(0, 0, 0, 0)
plot_save("anova-2x2-generic-blank.svg", width = width, height = height)
```

:::
