---
title: Problem Set 3
subtitle: Describing data

format: html
sidebar: recitation
navbar: false
execute: 
  echo: true
---

## Part 1: Computing descriptive statistics

### Central tendency

Suppose we have a set of values: `1, 2, 3, 3, 4, 5, 10`

There are several ways you could find the mean. One way would be to add the scores and divide by $N$, like you would on paper.

```{r}
(1 + 2 + 3 + 3 + 4 + 5 + 10) / 7
```

Of course, you don't want to have to type all those numbers and figure out $N$ yourself (especially if you were dealing with a much larger set of numbers). Better to have the scores as a named object:

```{r}
scores <- c(1, 2, 3, 3, 4, 5, 10)
```

Now life is much easier. The same mathematical approach can be implemented with the `sum()` and `length()` functions.

```{r}
sum(scores) / length(scores) # this is equivalent to Sigma X divided by N
```

Even better, R has built-in functions to easily find the mean and median of a set of scores.

```{r}
mean(scores)

median(scores)
```

You'd think there'd be one for finding the mode as well...

```{r}
mode(scores)
```

But that actually does something different.

Since the mode is a common statistic, though, somebody already wrote a function to compute it. It's in a package called `modeest`. The package contains the function `mfv1()`, for "most frequent value", which we can now use to get the mode. Remember, since it's an external package, we need to activate it with the `library()` function.

```{r}
library(modeest)
mfv1(scores)
```

Or equivalently, you can use the double-colon operator to refer to a `package::its_function()` without a separate `library()` call:

```{r}
modeest::mfv1(scores)
```

### Variability

There are functions to find the minimum and maximum value in a set of scores. The range is the difference between those:

```{r}
min(scores)
max(scores)
max(scores) - min(scores) # range
```

There is a built-in `range()` function. It returns the actual lowest and highest values as a collection. To find the difference between them, you could nest the range inside the `diff()` function.

```{r}
range(scores)
diff(range(scores))
```

R has a built-in functions for sample SD.

```{r}
sd(scores)
```

There's also a `var()` function to compute the variance. Remember that standard deviation is the square-root of variance, so if we square the SD it should give the same answer as `var()`...

```{r}
var(scores)
sd(scores)^2
```

Neato!

Note that these functions return *sample* variance & SD. There are no built-in functions for *population* variance & SD so if you wanted to find them you would have to use the mathematical procedure outlined in the lecture.

### Summarizing a data.frame column

The above instructions show how mathematical operations and functions can be used to compute summary statistics for a set of numbers. Most often, however, the set of numbers we're dealing with is a column in a data.frame.

```{r}
df <- data.frame(scores = c(1, 2, 3, 3, 4, 5, 10))
```

You can use the same approaches as above, just remember the `$` notation to refer to a data.frame column:

```{r}
mean(df$scores)
sd(df$scores)
```

However, `dplyr` has a powerful `summarize()` function that fits in nicely with the pipe operator.

```{r}
#| message: false

library(dplyr)

df |> 
  summarize(mean = mean(scores),
            median = median(scores),
            mode = modeest::mfv1(scores),
            min = min(scores),
            max = max(scores),
            range = max - min,
            sd = sd(scores))

```

In a single pipeline we produce all the summary statistics as a data.frame! Note that because I pipe `df` into the `summarize()` function, `scores` refers specifically to that column in the `df` data.frame object rather than the separate `scores` object I created previously. Note also that I defined the `min` and `max` summary columns, and I was able to refer to *those* new columns within the same single `mutate()` function to compute the `range`.


# Part 2. Comparing groups

### Reshaping data from wide to long format

The `dplyr::summarize()` approach becomes especially useful when we want to compare groups of scores, such as scores from different experimental groups in a study. In addition to `dplyr` I'll use another related package called `tidyr` here to change the structure of some data, making it easier to summarize and visualize (using `ggplot2`).

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
```


Suppose we have a data.frame with four columns; a participant ID and scores from experimental condition A, condition B, and condition C.

```{r}
demo_wide <- read.csv("data/3_demo.csv")

demo_wide
```

This is a very common data format. It's called "wide," because each variable gets its own column; the spreadsheet gets wider with each additional variable measured. However, an alternative data structure is "long"-formatted data. Rather than having multiple columns for each variable, there would be one column to identify the variable, and a second storing the value. Rather than getting wider, the spreadsheet gets longer, because each participant's responses take up multiple *rows* rather than multiple *columns*. Let's use `tidyr`'s `pivot_longer()` function to show what I mean.

```{r}
demo_long <- demo_wide |> 
  pivot_longer(-participant, 
               names_to = "condition",
               values_to = "score")

demo_long
```

Note that only the data columns should be pivoted, *not* the participant ID column; that's why the first argument `-participant`, excluded that column from the pivot operation. The result is that the new long-format data.frame has 27 rows in contrast to the original's 9. There are three columns: one for participant, with the value identifying each participant duplicated down the rows; a column named "condition" (thanks to the argument `names_to = "condition"`) containing the labels `condition_a`, `condition_b` and `condition_c`; and finally a column named `score` containing the recorded values associated with each participant and each condition.

### Describing data `.by` group

Why bother? Because now it is very easy to produce summary statistics for the whole dataset broken down into the different groups, using `summarize()`'s special `.by` argument to specify the grouping variable.

```{r}
demo_long |> 
  summarize(mean = mean(score),
            median = median(score),
            mode = modeest::mfv1(score),
            sd = sd(score),
            .by = condition)
```

### Visualizing grouped data

It is also a breeze to make graphs showing the distribution of the data grouped by condition. In this case, the column identifying which condition a score came from becomes another "aesthetic" in the `aes()` function–`color`, meaning the color of the lines and points. Note that depending on the nature of your data (its range and distribution) you may need to change the `binwidth` of the `geom_freqpoly()`. If you are also drawing a layer of points, you will need to also specify `stat = "bin"` and provide the same `binwidth` for that layer.

```{r}
demo_long |> 
  ggplot(aes(x = score, color = condition)) +
  geom_freqpoly(binwidth = 1) +
  geom_point(stat = "bin", binwidth = 1)
```

You can also take summarized data and pipe it straight into ggplot:

```{r}
demo_long |> 
  summarize(mean = mean(score),
            sd = sd(score),
            .by = condition) |> 
  ggplot(aes(x = condition, y = mean)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.3)
```

Note that I used `geom_col()` rather than `geom_bar()`, and `geom_errorbar()` required a couple of extra aesthetics to determine the min and max y positions, as well as an argument determining the `width` of the tops and bottoms of the errorbars.

